---
title: "Case study: verbal aggression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: verbal aggression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_data <- requireNamespace("psychotools", quietly = TRUE)
knitr::opts_chunk$set(eval = has_data)
library(rasch)
```

The verbal aggression data contain the responses of 316 students to 24
items (De Boeck and Wilson, 2004; Smits, De Boeck, and Vansteelandt, 2004).
Each item combines one of four frustrating situations, one of three possible
responses---cursing, scolding or shouting---and whether the respondent wants
to respond that way or would actually do so. Responses are `no`, `perhaps`
and `yes`, scored 0, 1 and 2.

This crossed item design makes the data useful for illustrating the linear
partial credit model (LPCM). Rather than estimate every item threshold
freely, the LPCM explains them from the characteristics used to construct the
items.

```{r data}
data("VerbalAggression", package = "psychotools")
responses <- VerbalAggression$resp
item_names <- colnames(responses)

item_design <- data.frame(
  item = item_names,
  situation = factor(substr(item_names, 1, 2), levels = paste0("S", 1:4)),
  mode = factor(ifelse(grepl("Want", item_names), "want", "do"),
                levels = c("do", "want")),
  behaviour = factor(
    sub("^S[1-4](Want|Do)", "", item_names),
    levels = c("Curse", "Scold", "Shout")
  )
)

head(item_design)
```

These are categorical item predictors. Person variables such as gender and
anger are not item predictors; they may instead be examined later through
targeting or differential item functioning.

## An additive model

We begin with additive effects for situation, mode and behaviour. The
reserved `threshold` factor distinguishes the transition from `no` to
`perhaps` from the transition from `perhaps` to `yes`. Its interaction with
mode allows that threshold structure to differ between wanting to act and
actually acting.

```{r additive}
fit_additive <- rasch_explanatory(
  responses,
  predictors = item_design,
  formula = ~ situation + mode + behaviour + threshold + mode:threshold
)

explanatory_test(fit_additive)
```

The Kent-adjusted comparison rejects this restriction
(\(\chi^2_{39} = 157.72\), \(p < .001\)).
The additive formulation does not capture all systematic differences among
the items.

## Interactions in the item design

The factorial construction also permits interactions among situation, mode
and behaviour. The following model includes every two-way interaction among
those characteristics and allows each characteristic to alter the response
threshold structure:

```{r interaction-model}
fit_explanatory <- rasch_explanatory(
  responses,
  predictors = item_design,
  formula = ~ (situation + mode + behaviour)^2 +
    threshold * (situation + mode + behaviour)
)

model_test <- explanatory_test(fit_explanatory)
model_test
```

This model estimates 24 explanatory coefficients in place of 47 free
threshold parameters. Its
Kent-adjusted comparison with the free calibration is not significant
(\(\chi^2_{23} = 26.30\), \(p = .287\)).
The explanatory structure therefore gives a materially smaller calibration
without a detected loss of fit in these data. It reproduces
`r round(100 * model_test$r_squared, 1)`% of the variation in the free
threshold calibration, and `r round(100 * model_test$r_squared_adj, 1)`%
after adjusting for the twenty-four coefficients doing the reproducing.

The threshold estimates from the explanatory and free calibrations are shown
below. Points close to the diagonal indicate where the predictor structure
reproduces the freely estimated threshold.

```{r threshold-comparison, fig.width=6, fig.height=5, fig.cap="Free and explanatory threshold estimates for the verbal aggression items."}
threshold_comparison <- merge(
  fit_explanatory$reference_fit$est$thr[c("item", "k", "tau")],
  fit_explanatory$est$thr[c("item", "k", "tau")],
  by = c("item", "k"),
  suffixes = c("_free", "_explanatory")
)

plot(
  threshold_comparison$tau_free,
  threshold_comparison$tau_explanatory,
  pch = c(1, 19)[threshold_comparison$k],
  xlab = "Free threshold estimate (logits)",
  ylab = "Explanatory threshold estimate (logits)"
)
abline(0, 1, lty = 2, col = "grey40")
legend("topleft", legend = c("First threshold", "Second threshold"),
       pch = c(1, 19), bty = "n")
```

The coefficient table is on the threshold-location scale. A positive
coefficient raises the relevant threshold and makes endorsement less likely;
a negative coefficient makes it more likely. Main effects refer to the
reference levels shown in `item_design`, and an interaction modifies the
corresponding main effect.

```{r coefficients}
effects <- fit_explanatory$est$coefficients
effects[effects$p_adj < .05,
        c("term", "estimate", "se", "z", "p", "p_adj")]
```

For example, the positive coefficients for `behaviourScold` and
`behaviourShout` indicate that these responses are harder to endorse than
cursing in the reference situation and mode. Their interactions show where
that difference changes. The threshold interactions describe changes in the
distance between `no`, `perhaps` and `yes`, rather than changes in the item
location as a whole.

## Checking fixed departures

The model comparison assesses the explanatory restrictions jointly.
`explanatory_diagnostics()` then checks whether any single item-location or
threshold-structure departure remains after the active model has been fitted.
Holm adjustment covers the complete set of candidate departures.

```{r departures}
departures <- explanatory_diagnostics(fit_explanatory)
head(departures, 10)
sum(departures$p_adj < .05, na.rm = TRUE)
```

No departure remains significant after adjustment, so there is no empirical
basis here for relaxing a particular item. If a substantively defensible
departure were found, `relax_explanatory()` would add it as a fixed effect and
repeat the calibration. The revised thresholds would then flow through to the
item estimates, person measures, residuals and subsequent analyses.

The sequence above is descriptive. In a confirmatory study, the interaction
structure should be specified from the item design before the responses are
examined.

## References

De Boeck, P., and Wilson, M. (Eds.). (2004). *Explanatory Item Response
Models: A Generalized Linear and Nonlinear Approach*. Springer.

Smits, D. J. M., De Boeck, P., and Vansteelandt, K. (2004). The inhibition of
verbally aggressive behaviour. *European Journal of Personality*, 18,
537--555. doi:10.1002/per.529.
