---
title: "Data structures for rasch models"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data structures for rasch models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r library}
library(rasch)
```

This vignette shows the data required by each model. The examples use plain
data frames so that the required columns are visible. Column names can differ;
name their roles in the fitting function.

## Rasch, partial credit and rating scale models

Use one row per person and one column per item. Scores begin at zero. A
dichotomous item is coded 0/1; an item with four categories is coded 0, 1, 2
or 3. Missing values may be `NA`; negative missing-value codes are also
accepted through `na_codes`.

```{r wide-data}
responses <- data.frame(
  person = c("P01", "P02", "P03", "P04"),
  group = c("control", "control", "treated", "treated"),
  I1 = c(0, 1, 1, 1),
  I2 = c(0, 1, 2, 2),
  I3 = c(1, 2, 3, NA)
)
responses
```

The identifier and person factors are not item responses. Name them
explicitly, or nominate the item columns.

```{r ordinary-fit, eval=FALSE}
fit <- rasch(
  responses,
  id = "person",
  factors = "group",
  items = c("I1", "I2", "I3"),
  model = "PCM"
)
```

Use `model = "RSM"` only when the items share one rating-scale threshold
structure. Items may have different maximum scores in the partial credit
model.

External weights are supplied after fitting. Item weights are a named numeric
vector. Set weights additionally need an item-to-set map, except for an EFRM
fit, which already contains one.

```{r person-weights, eval=FALSE}
item_weights <- c(I1 = 2, I2 = 1, I3 = 0.5)
weighted_person_estimates(fit, item_weights)

set_of <- c(I1 = "core", I2 = "core", I3 = "extension")
set_weights <- c(core = 2, extension = 1)
weighted_person_estimates(fit, set_weights, by = "set", sets = set_of)
```

Only relative weights matter. The weighted estimates are supplementary; they
do not replace the Rasch estimates used for fit, reliability, targeting or
DIF. In the application, use a CSV with `item,weight`, or
`item,set,weight` for set weights.

## Explanatory item and threshold models

The response data remain in the wide form above. A second data frame describes
the items. Item-level predictors have one row per item and an `item` column.

```{r item-predictors}
item_design <- data.frame(
  item = c("I1", "I2", "I3"),
  format = c("selected", "constructed", "constructed"),
  demand = c(0.2, 0.7, 1.1)
)
item_design
```

```{r explanatory-item-fit, eval=FALSE}
fit <- rasch_explanatory(
  responses,
  predictors = item_design,
  formula = ~ format + demand + format:demand,
  level = "item",
  id = "person",
  factors = "group",
  items = c("I1", "I2", "I3")
)
```

Categorical predictors should be factors, continuous predictors numeric, and
ordinal predictors ordered factors with their substantive order declared.
When a predictor varies between thresholds within an item, use one row per
threshold and include `item` and `threshold`.

```{r threshold-predictors}
threshold_design <- data.frame(
  item = rep(c("I2", "I3"), c(2, 3)),
  threshold = c(1, 2, 1, 2, 3),
  wording_load = c(0.2, 0.4, 0.1, 0.5, 0.9)
)
threshold_design
```

```{r explanatory-threshold-fit, eval=FALSE}
fit <- rasch_explanatory(
  responses,
  predictors = threshold_design,
  formula = ~ wording_load + threshold,
  level = "threshold",
  id = "person",
  factors = "group",
  items = c("I2", "I3")
)
```

## Multiple ratings

Long data use one row per observed rating. The required roles are person,
item, score and at least one facet. A rater, task or occasion can be a facet.
A person-group variable is instead supplied through `factors`.

```{r mfrm-long}
ratings <- data.frame(
  person = c("P01", "P01", "P02", "P02"),
  item = c("Essay1", "Essay2", "Essay1", "Essay2"),
  rater = c("R1", "R2", "R2", "R1"),
  occasion = c("first", "first", "first", "first"),
  score = c(2, 3, 1, 2)
)
ratings
```

```{r mfrm-fit, eval=FALSE}
fit <- rasch_mfrm(
  ratings,
  person = "person",
  item = "item",
  score = "score",
  facets = c("rater", "occasion")
)
```

Wide multiple-ratings data use one row per person-by-facet combination and
one score column per item.

```{r mfrm-wide, eval=FALSE}
wide_ratings <- data.frame(
  person = c("P01", "P01", "P02", "P02"),
  rater = c("R1", "R2", "R1", "R2"),
  Essay1 = c(2, 3, 1, 2),
  Essay2 = c(3, 2, 2, 2)
)

fit <- rasch_mfrm(
  wide_ratings,
  person = "person",
  facets = "rater",
  items = c("Essay1", "Essay2")
)
```

Long data are usually easier to inspect when a facet varies within items.

## Extended Frames

EFRM response data use one row per person and one column per item, together
with a person-group column. A separate named map assigns each item to exactly
one item set.

```{r efrm-structure, eval=FALSE}
frame_data <- data.frame(
  person = paste0("P", 1:6),
  group = rep(c("A", "B"), each = 3),
  S1I1 = c(0, 1, 1, 0, 1, 1),
  S1I2 = c(0, 0, 1, 0, 1, 1),
  S2I1 = c(0, 1, 1, 0, 0, 1),
  S2I2 = c(0, 1, 1, 0, 1, 1)
)
item_sets <- list(
  set1 = c("S1I1", "S1I2"),
  set2 = c("S2I1", "S2I2")
)

fit <- rasch_efrm(
  frame_data,
  item_sets = item_sets,
  groups = "group",
  id = "person"
)
```

Persons must connect the item sets, and common items across person groups must
identify the group units. Crossed person-group factors can be supplied as
several column names in `groups`.

## Comparative Judgement

Use one row per comparison. `object_a` and `object_b` identify the pair. For a
dichotomous comparison, `winner` contains one of those two object names.

```{r cj-data}
comparisons <- data.frame(
  object_a = c("A", "A", "B", "A"),
  object_b = c("B", "C", "C", "C"),
  winner = c("A", "C", "B", "A"),
  judge = c("J1", "J1", "J2", "J2")
)
comparisons
```

```{r cj-fit, eval=FALSE}
fit <- btl(
  comparisons,
  object_a = "object_a",
  object_b = "object_b",
  winner = "winner",
  judge = "judge"
)
```

For ordered comparisons, replace `winner` with an ordered response column
whose categories run from preference for object B to preference for object A.
A frequency column can represent repeated identical rows. A judgement-order
column is needed to estimate exposure or carry-over dependence.

Explanatory Comparative Judgement adds one metadata row per object.

```{r cj-explanatory, eval=FALSE}
object_design <- data.frame(
  object = c("A", "B", "C"),
  genre = factor(c("essay", "essay", "report")),
  length = c(800, 950, 700)
)

fit <- btl_explanatory(
  comparisons,
  predictors = object_design,
  formula = ~ genre + length,
  object_a = "object_a",
  object_b = "object_b",
  winner = "winner",
  judge = "judge"
)
```

## Extended Frames for Comparative Judgement

The comparison rows additionally identify each judge's panel. A named list
assigns every object to one object set. Both within-set and cross-set
comparisons are required to identify set units and origins.

```{r cj-frames, eval=FALSE}
object_sets <- list(
  set1 = c("S1A", "S1B", "S1C"),
  set2 = c("S2A", "S2B", "S2C")
)

fit <- btl_efrm(
  frame_comparisons,
  object_a = "object_a",
  object_b = "object_b",
  winner = "winner",
  judge = "judge",
  panels = "panel",
  object_sets = object_sets
)
```

## Simulated data

The **Simulate** page in the Shiny application generates every structure
above, assigns its roles and retains the generating values. The corresponding
R functions are `simulate_rasch()`, `simulate_btl()`, `simulate_mfrm()`,
`simulate_efrm()` and `simulate_btl_efrm()`. Explanatory simulations add the
item or object metadata used to generate the locations. Each simulator stores
its parameter values and planted departures in `attr(data, "truth")`. The app
can download the data as a CSV or as a bundle containing the generating call,
true values and explanatory metadata.

```{r simulation-map}
d <- simulate_efrm(n_per_group = 100, items_per_set = 5, seed = 4)
names(attr(d, "truth"))
```
