---
title: "DIF with several person factors and repeated measures"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DIF with several person factors and repeated measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r library}
library(rasch)
```

## Design the analysis around the person

When several person factors are relevant, they should enter one model. Testing
them one at a time can assign shared variation according to the order of the
analyses. Repeated observations require a further distinction: group is a
between-person factor, whereas occasion varies within person.

The following dataset has two observations per person. Item I03 has a group
shift and item I06 has an occasion shift.

```{r data}
set.seed(21)
N <- 240
difficulty <- seq(-1.5, 1.5, length.out = 8)
theta <- rnorm(N)
group <- rep(c("A", "B"), each = N / 2)

make_wave <- function(occasion_shift) {
  shift <- matrix(0, N, 8)
  shift[group == "B", 3] <- 0.8
  shift[, 6] <- occasion_shift
  matrix(rbinom(N * 8, 1,
                plogis(outer(theta, difficulty, "-") - shift)), N, 8)
}

X <- rbind(make_wave(0), make_wave(0.9))
colnames(X) <- sprintf("I%02d", 1:8)
dat <- data.frame(
  pid = rep(sprintf("P%03d", seq_len(N)), 2),
  X,
  group = rep(group, 2),
  occasion = rep(c("T1", "T2"), each = N)
)
```

## Fit once and test both factors

The repeated person identifier is carried into the fit. `dif_anova` detects
occasion as within-person; it can also be declared explicitly. Persons, not
stacked rows, are the units of analysis. Between-person tests use Type II sums
of squares. Within-person tests use person-level contrasts, with a
Greenhouse--Geisser correction when a factor has more than two levels.
This joint multi-factor, mixed-design implementation extends the conventional
single-factor residual analysis of variance of Andrich and Marais (2019).
Its F references remain approximate,
so consequential or sparse designs should also be checked by simulation at
their observed cell sizes and missingness pattern.

```{r analysis}
fit <- rasch(dat, id = "pid", factors = c("group", "occasion"),
             items = sprintf("I%02d", 1:8))
da <- dif_anova(fit, within = "occasion")
da$summary
```

At this seed the two planted effects are flagged clearly: I03 by group and
I06 by occasion. A third, weaker occasion flag appears on I05 (adjusted p
just under 0.05): no effect was planted there, and a borderline extra flag
of this kind is exactly what a 5 per cent criterion over many item-by-term
tests produces from time to time -- read isolated marginal flags with the
magnitude estimates below before acting on them.

The multiplicity adjustment is applied across items separately within each
term. Uniform DIF is a factor effect that is stable over the trait; a
factor-by-class-interval effect is non-uniform DIF. If person-factor
interactions are substantively required, use `effects = "factorial"` and
interpret a significant higher-order term before its component main effects.

## Quantify the departure

ANOVA identifies evidence against invariance; it does not state the size of
the departure in logits. `dif_size` resolves an item by a between-person
factor. `dif_contrasts` is the corresponding planned-contrast approach and
uses person-level differencing for within-person questions.

```{r magnitude}
dif_size(fit, "I03", by = "group")
dc <- dif_contrasts(fit, items = c("I03", "I06"), within = "occasion",
                    id = fit$person$id)
dc$table
```

For repeated-person contrasts, significance comes from person-level residual
contrast scores. The resolved logit difference remains the magnitude, but its
row-independent calibration covariance is not a repeated-measures standard
error; the package therefore withholds the logit SE and interval in this case.

A statistical flag should be considered with the logit magnitude, targeting,
item content, and the intended use of the scale. Resolving an item changes the
measurement model and should follow a substantive account of why the item is
not invariant.
