---
title: "DIF with several person factors and repeated measures"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DIF with several person factors and repeated measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r library}
library(rasch)
```

## Design the analysis around the person

Differential item functioning is a violation of the Rasch model's requirement
of invariant comparison: item locations should not depend on which persons
respond (Rasch 1961). When several person factors are relevant, they should
enter one model. Repeated observations require a further distinction: group is
a between-person factor, whereas occasion varies within person.

For a factor $G$, class interval $C$, and standardised residual $z$, the
single-factor model is

$$
z=\mu+G+C+G\mathbin{:}C+\varepsilon.
$$

The $G$ term tests uniform DIF. The $G\mathbin{:}C$ term tests non-uniform DIF.
With several factors, `dif_anova` fits their terms jointly and uses Type II
sums of squares.

The following dataset has two observations per person. Item I03 has a group
shift and item I06 has an occasion shift.

```{r data}
set.seed(21)
N <- 240
difficulty <- seq(-1.5, 1.5, length.out = 8)
theta <- rnorm(N)
group <- rep(c("A", "B"), each = N / 2)

make_wave <- function(occasion_shift) {
  shift <- matrix(0, N, 8)
  shift[group == "B", 3] <- 0.8
  shift[, 6] <- occasion_shift
  matrix(rbinom(N * 8, 1,
                plogis(outer(theta, difficulty, "-") - shift)), N, 8)
}

X <- rbind(make_wave(0), make_wave(0.9))
colnames(X) <- sprintf("I%02d", 1:8)
dat <- data.frame(
  pid = rep(sprintf("P%03d", seq_len(N)), 2),
  X,
  group = rep(group, 2),
  occasion = rep(c("T1", "T2"), each = N)
)
```

## Fit once and test both factors

The repeated person identifier is carried into the fit. `dif_anova` detects
occasion as within-person; it can also be declared explicitly. Persons, not
stacked rows, are the units of analysis. Between-person tests use Type II sums
of squares. Within-person tests use person-level contrasts, with a
Greenhouse--Geisser correction when a factor has more than two levels.
This mixed-design analysis extends the single-factor residual analysis of
variance described by Andrich and Marais (2019). Its F references are
large-sample approximations.

```{r analysis}
fit <- rasch(dat, id = "pid", factors = c("group", "occasion"),
             items = sprintf("I%02d", 1:8))
da <- dif_anova(fit, within = "occasion", sizes = TRUE)
da$summary
```

The multiplicity adjustment is applied across items separately within each
term. Uniform DIF is a factor effect that is stable over the trait; a
factor-by-class-interval effect is non-uniform DIF. If person-factor
interactions are substantively required, use `effects = "factorial"` and
interpret a significant higher-order term before its component main effects.
Read adjusted probabilities with effect sizes before changing an item.

## Quantify the departure

ANOVA identifies evidence against invariance; it does not state the size of
the departure in logits. `dif_size` resolves an item by a between-person
factor. `dif_contrasts` provides planned contrasts and uses person-level
differencing for within-person questions.

```{r magnitude}
dif_size(fit, "I03", by = "group")
dc <- dif_contrasts(fit, items = c("I03", "I06"), within = "occasion")
dc$table
da$posthoc
```

`dif_posthoc()` is the general follow-up for a significant term. A main
effect with more than two levels is reported as pairwise marginal differences
over the other fitted factors. An interaction is reported as a
difference-in-differences, or its higher-order counterpart. These comparisons
use the joint covariance of the resolved item locations and Holm adjustment
over the stated family. They are preferable to an ordinary Tukey comparison
of residual means because the result is on the logit scale and respects the
factor structure used in the DIF analysis.

For repeated-person contrasts, significance comes from person-level residual
contrast scores with the same design-cell weights as the resolved estimate.
Other fitted factors are averaged equally over their complete cells, including
when their sample sizes differ, and the independent between-person cells use a
Welch--Satterthwaite reference. The resolved logit difference remains the
magnitude, but its row-independent calibration covariance is not a
repeated-measures standard error; the package therefore withholds the logit SE
and interval in this case. The fitted person identifier is used unless `id` is
supplied explicitly.

For a many-facet fit, follow-ups may name the underlying item; its virtual
facet cells are pooled with common weights so facet severity cancels from the
group contrast. Extended-frame fits support the residual ANOVA for factors
outside the frame definition, but not an ordinary resolved-item magnitude:
that refit would discard the fitted frame units.

A statistical flag should be considered with the logit magnitude, targeting,
item content, and the intended use of the scale. Resolving an item changes the
measurement model and should follow a substantive account of why the item is
not invariant.

## References

Andrich, D., and Marais, I. (2019). *A Course in Rasch Measurement Theory:
Measuring in the Educational, Social and Health Sciences*. Springer.

Rasch, G. (1961). On general laws and the meaning of measurement in
psychology. In *Proceedings of the Fourth Berkeley Symposium on Mathematical
Statistics and Probability* (Vol. 4, pp. 321--333). Berkeley: University of
California Press.
