---
title: "DIF with several person factors and repeated measures"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DIF with several person factors and repeated measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r library}
library(rasch)
```

## Design the analysis around the person

Differential item functioning is a violation of the Rasch model's requirement
of invariant comparison: item locations should not depend on which persons
respond (Rasch 1961). When several person factors are relevant, they should
enter one model. Repeated observations require a further distinction: group is
a between-person factor, whereas occasion varies within person.

For a factor $G$, class interval $C$, and standardised residual $z$, the
single-factor model is

$$
z=\mu+G+C+G\mathbin{:}C+\varepsilon.
$$

The $G$ term tests uniform DIF. The $G\mathbin{:}C$ term tests non-uniform DIF.
With several factors, `dif_anova` fits their terms jointly and uses Type II
sums of squares.

The following dataset has two observations per person. Item I03 has a group
shift and item I06 has an occasion shift.

```{r data}
set.seed(21)
N <- 240
difficulty <- seq(-1.5, 1.5, length.out = 8)
theta <- rnorm(N)
group <- rep(c("A", "B"), each = N / 2)

make_wave <- function(occasion_shift) {
  shift <- matrix(0, N, 8)
  shift[group == "B", 3] <- 0.8
  shift[, 6] <- occasion_shift
  matrix(rbinom(N * 8, 1,
                plogis(outer(theta, difficulty, "-") - shift)), N, 8)
}

X <- rbind(make_wave(0), make_wave(0.9))
colnames(X) <- sprintf("I%02d", 1:8)
dat <- data.frame(
  pid = rep(sprintf("P%03d", seq_len(N)), 2),
  X,
  group = rep(group, 2),
  occasion = rep(c("T1", "T2"), each = N)
)
```

## Fit once and test both factors

The repeated person identifier is carried into the fit. `dif_anova` detects
occasion as within-person; it can also be declared explicitly. Persons, not
stacked rows, are the units of analysis. Between-person tests use Type II sums
of squares. Within-person tests use person-level contrasts, with a
Greenhouse--Geisser correction when a factor has more than two levels.
This mixed-design analysis extends the single-factor residual analysis of
variance described by Andrich and Marais (2019). Its F references are
large-sample approximations.

```{r analysis}
fit <- rasch(dat, id = "pid", factors = c("group", "occasion"),
             items = sprintf("I%02d", 1:8))
da <- dif_anova(fit, within = "occasion")
da$summary
```

The planted effects are I03 by group and I06 by occasion, and both are
flagged. This run also flags I05 by occasion, where nothing was planted:
with eight items tested against every term, an occasional spurious flag is
expected at any fixed level, which is why adjusted probabilities and effect
sizes should be read together before an item is acted on. The `dif_size`
step below shows the planted effects are large where the spurious flag is
not.

The multiplicity adjustment is applied across items separately within each
term. Uniform DIF is a factor effect that is stable over the trait; a
factor-by-class-interval effect is non-uniform DIF. If person-factor
interactions are substantively required, use `effects = "factorial"` and
interpret a significant higher-order term before its component main effects.

## Quantify the departure

ANOVA identifies evidence against invariance; it does not state the size of
the departure in logits. `dif_size` resolves an item by a between-person
factor. `dif_contrasts` is the corresponding planned-contrast approach and
uses person-level differencing for within-person questions.

```{r magnitude}
dif_size(fit, "I03", by = "group")
dc <- dif_contrasts(fit, items = c("I03", "I06"), within = "occasion",
                    id = fit$person$id)
dc$table
```

For repeated-person contrasts, significance comes from person-level residual
contrast scores. The resolved logit difference remains the magnitude, but its
row-independent calibration covariance is not a repeated-measures standard
error; the package therefore withholds the logit SE and interval in this case.

A statistical flag should be considered with the logit magnitude, targeting,
item content, and the intended use of the scale. Resolving an item changes the
measurement model and should follow a substantive account of why the item is
not invariant.

## References

Andrich, D., and Hagquist, C. (2012). Real and artificial differential item
functioning. *Journal of Educational and Behavioral Statistics*, 37(3),
387--416.

Andrich, D., and Hagquist, C. (2015). Real and artificial differential item
functioning in polytomous items. *Educational and Psychological Measurement*,
75(2), 185--207.

Andrich, D., and Marais, I. (2019). *A Course in Rasch Measurement Theory:
Measuring in the Educational, Social and Health Sciences*. Springer.

Hagquist, C., and Andrich, D. (2017). Recent advances in analysis of
differential item functioning in health research using the Rasch model.
*Health and Quality of Life Outcomes*, 15, 181.

Maxwell, S. E., and Delaney, H. D. (2004). *Designing Experiments and
Analyzing Data: A Model Comparison Perspective* (2nd ed.). Lawrence Erlbaum.

Rasch, G. (1961). On general laws and the meaning of measurement in
psychology. In *Proceedings of the Fourth Berkeley Symposium on Mathematical
Statistics and Probability* (Vol. 4, pp. 321--333). Berkeley: University of
California Press.
