---
title: "The extended frame of reference model"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The extended frame of reference model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r library}
library(rasch)
```

## Frames and units

Rasch's criterion of invariant comparison holds within a specified frame of
reference (Rasch 1961). The extended frame of reference model (Humphry 2005;
Humphry and Andrich 2008) makes the frame explicit: it allows the unit to
differ across linked item-set by person-group frames. Within each frame, the
partial credit model holds in that frame's natural unit; dichotomous items
are its one-threshold special case, and the example below uses them.

For item $i$ in set $s$ and person $n$ in group $g$,

$$
P(X_{ni}=x)=
\frac{\exp\left\{\rho_{sg}\left[x\theta_n-
\sum_{k=1}^{x}\delta_{ik}\right]\right\}}
{\sum_{y=0}^{m_i}\exp\left\{\rho_{sg}\left[y\theta_n-
\sum_{k=1}^{y}\delta_{ik}\right]\right\}},
\qquad \rho_{sg}=\alpha_s\phi_g.
$$

Here $\alpha_s$ is the item-set unit and $\phi_g$ is the person-group unit.
A frame-defining factor sets the unit and cannot also be tested for DIF:
the factor takes a single level among the persons responding within any
frame, so no within-frame comparison remains on which to test it. This is
why `dif_anova()` refuses such a factor, and why the fitted model cannot
check its own assumption: it holds each item at one location across frames
by construction, so the per-frame estimates it reports agree because the
model made them agree. Testing the assumption means stepping outside the
model, which is what `frame_invariance()` does -- see below.

The model should be used when a frame-dependent unit is part of the
substantive measurement account, not simply because an equal-unit model fits
poorly.

```{r simulate}
d <- simulate_efrm(n_per_group = 500, items_per_set = 12,
                   set_unit_ratio = 1.30, group_unit_ratio = 1.10,
                   seed = 25)
truth <- attr(d, "truth")
```

The simulator defaults to dichotomous items;
`simulate_efrm(n_categories = 4)` generates partial credit items within
frames instead, which carry more linking information per item.

## Fit and inspect the links

```{r fit}
fit <- rasch_efrm(
  d,
  item_sets = truth$item_sets,
  groups = "group",
  id = "id",
  boot_reps = 60
)
fit
```

`phi_table` reports person-group units, `alpha_table` reports item-set units,
and `set_table` reports the linked set locations. Units are positive and are
most naturally compared as ratios. The linking tables and notes should be read
before interpreting common-unit item or person estimates.

Both recovered units sit close to the planted 1.30 and 1.10 here, but read
that as one draw rather than as the estimator's precision. Units are ratios
of estimated dispersions and carry real sampling error: at this design
repeated draws give set-unit ratios with a standard deviation of about
0.055 and group-unit ratios about 0.038, so analyses routinely land several
per cent from the generating value in either direction. This example uses a
seed whose draw falls near the centre of that distribution; a different
seed would land elsewhere within it.

The two units are estimated differently and behave differently. Group units
come from the within-frame conditional calibration: unbiased, with
well-calibrated intervals (no detectable bias over 60 replicates at this
design, 95% coverage). Set units come from the linking step and additionally
carry a small fixed-design offset, around one per cent at these set
lengths.

Judge a recovered unit against its standard error, then, and never against
a single alternative value alone. Whether the *estimator* is on target is a
question about repeated draws, not about one:

```{r recovery-check, eval = FALSE}
ratios <- vapply(seq_len(20), function(r) {
  dd <- simulate_efrm(n_per_group = 500, items_per_set = 12,
                      set_unit_ratio = 1.30, group_unit_ratio = 1.10,
                      seed = 91000 + r)
  ff <- rasch_efrm(dd, item_sets = attr(dd, "truth")$item_sets,
                   groups = "group", id = "id", boot_reps = 0)
  c(alpha = max(ff$alpha_table$alpha) / min(ff$alpha_table$alpha),
    phi = max(ff$phi_table$phi) / min(ff$phi_table$phi))
}, c(alpha = 0, phi = 0))
rowMeans(ratios)   # 1.304 and 1.107 against the planted 1.30 and 1.10
apply(ratios, 1, sd)
```

Longer item sets, not larger samples, are what tighten a set unit: the
precision of the set-linking step is limited by the number of items in each
set once a few hundred linking persons are available.

```{r tables}
fit$phi_table
fit$alpha_table
fit$set_table
fit$linking
```

Group units are estimated from within-frame pairwise conditional calibrations.
Set units and locations require persons common to the linked sets. A
disconnected design cannot establish a common unit.

The set-linking step uses a method-of-moments estimator based on the
true-score variance-ratio argument in Humphry (2005), with the true-score
variance recovered by a truncated-score-moment correction: the mean and
variance of the score-to-measure map over the non-extreme scores are exact
model functions of the person location, and their expectations are
estimated through score weights whose validity does not depend on the
person distribution. This matters at any realistic length: the classical
"observed variance minus mean squared standard error" construction
under-recovers the true-score variance by more than half at eight
dichotomous items per set, and its errors do not fade with length -- in
simulation it biased recovered unit ratios by about +5 per cent at eight
items and -2 to -4 per cent at sixteen to thirty-two, never converging in
the tested range, and it fails entirely below seven items. The corrected
estimator is unbiased at every tested set length, unit ratio (1 to 2),
and person distribution (skewness to 2.8, where a normal-population
marginal maximum likelihood anchor drifts), and agrees with independent
external anchors. Its remaining fixed-design offset is small -- about 0.4
per cent on the unit ratio at eight dichotomous items per set, decaying
with set length -- and intervals stay calibrated to at least five
thousand linking persons; beyond that, accuracy is bounded by the set
length, not the sample. Each linking person's pattern must span a score
range of at least four within a set (four dichotomous items; six or more
are recommended), and because the correction computes score distributions
from the fitted within-frame model, model violations within a set bias
the recovered units roughly in proportion. The estimator is distinct from
the likelihood proposed in section 5.3 of that thesis. The crossed,
multigroup, and polytomous forms are extensions implemented in this
package.

## Uncertainty and comparison with equal units

The default hybrid standard errors combine the pairwise sandwich covariance,
a person bootstrap for the set-linking stage, and delta-method propagation.
Across linking samples from 250 to 10,000 persons the reported standard
errors track the empirical sampling variability as both shrink with the
square root of the sample.
The example uses 60 replicates to keep the vignette quick; a final analysis
should use enough replicates to stabilise its reported standard errors. Set
`se_method = "bootstrap"` to refit the complete pipeline on each person
resample.

```{r comparison}
fit$frames
fit$efrm_vs_rasch
```

The raw equal-unit composite-likelihood difference is descriptive. The omnibus
Wald tests in `efrm_vs_rasch$unit_omnibus` test the unit families; individual
unit contrasts are Holm-adjusted follow-ups.

## Testing the invariance the model assumes

`frame_invariance()` calibrates each frame separately, puts the item
locations on the common scale by dividing by that frame's unit, and compares
them item by item. It reports two comparisons, and they are not equally
sensitive: a location difference is found about 97 per cent of the time at
500 persons per frame, a difference in discrimination about 16 per cent, and
the two reach comparable power only near 2,000. A clean result at a few
hundred persons per frame is therefore much stronger evidence against
differential item functioning than against differential discrimination.

```{r invariance, eval = FALSE}
inv <- frame_invariance(fit)
inv$summary                    # rmsd against rmse, per set and frame pair
inv$locations
```

Which items to flag is a screening decision, not a confirmatory one, and
`adjust` chooses. Holm across eight to ten items is right for reporting that
an item differs and wrong for deciding which items to look at, where it
costs between 20 and 60 points of sensitivity. Both probabilities are
reported either way, so the choice changes only which rows are flagged.

```{r screen, eval = FALSE}
frame_invariance(fit, adjust = "none")   # screen
```

An item the test flags is removed with `drop_items()`, which refits and
returns an ordinary fit of the same class. Rank the items by `abs(fit_resid)`
rather than applying a fixed cut-off -- a cut states detectability rather
than magnitude, so it flags more items the more persons there are -- then
drop the largest departure, refit, and stop when the units no longer differ.
Dropping is a complete cure where the item is found; what limits the repair
is detection, not removal.

```{r repair, eval = FALSE}
fit2 <- drop_items(fit, "I07")
fit2$efrm_vs_rasch$unit_tests            # has the difference gone?
```

## Worked analyses on real data

Three case studies apply the frame of reference models to real datasets and
ship with the package. The two wording studies are best read together: they
find a set-unit difference of almost the same size and it means something
different in each. In the self-esteem data the difference is carried by one
ambivalent item and vanishes when that item goes; in the height data it
survives every single-item removal, and yet leaves the person ordering
essentially unchanged against a criterion collected outside the inventory.
Neither conclusion is available from the unit test alone.

```{r casestudies, eval = FALSE}
# wording effects as item-set units in a self-esteem scale
file.edit(system.file("casestudies", "wording_units_selfesteem.R",
                      package = "rasch"))
# the same question in a balanced inventory, checked against a criterion
file.edit(system.file("casestudies", "wording_units_height.R",
                      package = "rasch"))
# panel and bloc frames in paired comparisons between political parties
file.edit(system.file("casestudies", "party_blocs_crisis.R",
                      package = "rasch"))
```

## References

Humphry, S. M. (2005). *Maintaining a Common Arbitrary Unit in Social
Measurement*. PhD thesis, Murdoch University.

Humphry, S. M., and Andrich, D. (2008). Understanding the unit in the Rasch
model. *Journal of Applied Measurement*, 9(3), 249--264.

Rasch, G. (1961). On general laws and the meaning of measurement in
psychology. In *Proceedings of the Fourth Berkeley Symposium on Mathematical
Statistics and Probability* (Vol. 4, pp. 321--333). Berkeley: University of
California Press.
