---
title: "The extended frame of reference model"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The extended frame of reference model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r library}
library(rasch)
```

## Frames and units

Rasch's criterion of invariant comparison holds within a specified frame of
reference (Rasch 1961). The extended frame of reference model (Humphry 2005;
Humphry and Andrich 2008) makes the frame explicit: it allows the unit to
differ across linked item-set by person-group frames. Within each frame, the
partial credit model holds in that frame's natural unit.

For item $i$ in set $s$ and person $n$ in group $g$,

$$
P(X_{ni}=x)=
\frac{\exp\left\{\rho_{sg}\left[x\theta_n-
\sum_{k=1}^{x}\delta_{ik}\right]\right\}}
{\sum_{y=0}^{m_i}\exp\left\{\rho_{sg}\left[y\theta_n-
\sum_{k=1}^{y}\delta_{ik}\right]\right\}},
\qquad \rho_{sg}=\alpha_s\phi_g.
$$

Here $\alpha_s$ is the item-set unit and $\phi_g$ is the person-group unit.
A frame-defining factor sets the unit and cannot also be tested for DIF:
the factor takes a single level among the persons responding within any
frame, so no within-frame comparison remains on which to test it.

The model should be used when a frame-dependent unit is part of the
substantive measurement account, not simply because an equal-unit model fits
poorly.

```{r simulate}
d <- simulate_efrm(n_per_group = 300, items_per_set = 8,
                   set_unit_ratio = 1.30, group_unit_ratio = 1.10,
                   seed = 12)
truth <- attr(d, "truth")
```

## Fit and inspect the links

```{r fit}
fit <- rasch_efrm(
  d,
  item_sets = truth$item_sets,
  groups = "group",
  id = "id",
  boot_reps = 60
)
fit
```

`phi_table` reports person-group units, `alpha_table` reports item-set units,
and `set_table` reports the linked set locations. Units are positive and are
most naturally compared as ratios. The linking tables and notes should be read
before interpreting common-unit item or person estimates.

```{r tables}
fit$phi_table
fit$alpha_table
fit$set_table
fit$linking
```

Group units are estimated from within-frame pairwise conditional calibrations.
Set units and locations require persons common to the linked sets. A
disconnected design cannot establish a common unit.

The set-linking step uses a method-of-moments estimator based on the
true-score variance-ratio argument in Humphry (2005), with the true-score
variance recovered by a truncated-score-moment correction: the mean and
variance of the score-to-measure map over the non-extreme scores are exact
model functions of the person location, and their expectations are
estimated through score weights that are unbiased for any person
distribution. On short tests this matters -- the naive
"observed variance minus mean squared standard error" construction
under-recovers the true-score variance by more than half at eight
dichotomous items per set and biased recovered unit ratios upward by
about five per cent, a bias confirmed against an external anchor and
removed by the correction. The estimator is distinct from the likelihood
proposed in section 5.3 of that thesis. The crossed, multigroup, and
polytomous forms are extensions implemented in this package.

## Uncertainty and comparison with equal units

The default hybrid standard errors combine the pairwise sandwich covariance,
a person bootstrap for the set-linking stage, and delta-method propagation.
The example uses 60 replicates to keep the vignette quick; a final analysis
should use enough replicates to stabilise its reported standard errors. Set
`se_method = "bootstrap"` to refit the complete pipeline on each person
resample.

```{r comparison}
fit$frames
fit$efrm_vs_rasch
```

The raw equal-unit composite-likelihood difference is descriptive. The omnibus
Wald tests in `efrm_vs_rasch$unit_omnibus` test the unit families; individual
unit contrasts are Holm-adjusted follow-ups.

## References

Humphry, S. M. (2005). *Maintaining a Common Arbitrary Unit in Social
Measurement*. PhD thesis, Murdoch University.

Humphry, S. M., and Andrich, D. (2008). Understanding the unit in the Rasch
model. *Journal of Applied Measurement*, 9(3), 249--264.

Rasch, G. (1961). On general laws and the meaning of measurement in
psychology. In *Proceedings of the Fourth Berkeley Symposium on Mathematical
Statistics and Probability* (Vol. 4, pp. 321--333). Berkeley: University of
California Press.
