---
title: "The extended frame of reference model"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The extended frame of reference model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r library}
library(rasch)
```

## Frames and units

Rasch's criterion of invariant comparison holds within a specified frame of
reference (Rasch 1961). The extended frame of reference model (Humphry 2005;
Humphry and Andrich 2008) makes the frame explicit: it allows the unit to
differ across linked item-set by person-group frames. Within each frame, the partial credit model
holds in that frame's natural unit. This is a different claim from ordinary
DIF: a frame-defining factor sets the scale unit and cannot also be tested as
a separate DIF factor on the frame's own virtual items.

The model should be used when a frame-dependent unit is part of the
substantive measurement account, not simply because an equal-unit model fits
poorly.

```{r simulate}
d <- simulate_efrm(n_per_group = 300, items_per_set = 8,
                   set_unit_ratio = 1.30, group_unit_ratio = 1.10,
                   seed = 12)
truth <- attr(d, "truth")
```

## Fit and inspect the links

```{r fit}
fit <- rasch_efrm(
  d,
  item_sets = truth$item_sets,
  groups = "group",
  id = "id",
  boot_reps = 60
)
fit
```

`phi_table` reports person-group units, `alpha_table` reports item-set units,
and `set_table` reports the linked set locations. Units are positive and are
most naturally compared as ratios. The linking tables and notes should be read
before interpreting common-unit item or person estimates.

```{r tables}
fit$phi_table
fit$alpha_table
fit$set_table
fit$linking
```

The group units are obtained from within-frame pairwise conditional
calibrations. Set units and locations require persons common to the linked
sets. A disconnected or weakly linked design cannot establish a common unit;
increasing the number of responses within an isolated frame does not repair
the missing link.

The person-side set-linking step is an error-corrected method-of-moments
implementation based on the true-score variance-ratio argument in Humphry
(2005), rather than the separate likelihood proposal in section 5.3 of that
thesis. The crossed, multigroup and polytomous implementation is an
experimental extension. Its sampling behaviour should be checked for the
intended design, preferably with the full person bootstrap. At this
vignette's design and sample size, for example, the recovered set-unit
ratio runs a modest few per cent above the planted value across
replications -- a finite-sample bias of the moments-based link, not
sampling noise alone -- which is exactly the kind of behaviour such a
check reveals before the estimates are interpreted substantively.

## Uncertainty and comparison with equal units

The default hybrid standard errors combine the pairwise sandwich covariance,
a person bootstrap for the set-linking stage, and delta-method propagation.
The example uses 60 replicates to keep the vignette quick; a final analysis
should use enough replicates to stabilise its reported standard errors. Set
`se_method = "bootstrap"` to refit the complete pipeline on each person
resample.

```{r comparison}
fit$frames
fit$efrm_vs_rasch
```

The raw equal-unit composite-likelihood difference is descriptive. Inference
is carried by the omnibus Wald tests in `efrm_vs_rasch$unit_omnibus`;
individual unit contrasts are exploratory and Holm-adjusted. None of these is
a substitute for checking frame-level fit, targeting, link strength, and the
defensibility of the frame interpretation.

## References

Andrich, D., and Marais, I. (2019). *A Course in Rasch Measurement Theory:
Measuring in the Educational, Social and Health Sciences*. Springer.

Humphry, S. M. (2005). *Maintaining a Common Arbitrary Unit in Social
Measurement*. PhD thesis, Murdoch University.

Humphry, S. M., and Andrich, D. (2008). Understanding the unit in the Rasch
model. *Journal of Applied Measurement*, 9(3), 249--264.

Rasch, G. (1960). *Probabilistic Models for Some Intelligence and Attainment
Tests*. Copenhagen: Danish Institute for Educational Research. (Expanded
edition, 1980, Chicago: University of Chicago Press.)

Rasch, G. (1961). On general laws and the meaning of measurement in
psychology. In *Proceedings of the Fourth Berkeley Symposium on Mathematical
Statistics and Probability* (Vol. 4, pp. 321--333). Berkeley: University of
California Press.
