---
title: "The extended frame of reference model"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The extended frame of reference model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r library}
library(rasch)
```

## Frames and units

Rasch's criterion of invariant comparison holds within a specified frame of
reference (Rasch 1961). The extended frame of reference model (Humphry 2005;
Humphry and Andrich 2008) makes the frame explicit: it allows the unit to
differ across linked item-set by person-group frames. Within each frame, the
partial credit model holds in that frame's natural unit.

For item $i$ in set $s$ and person $n$ in group $g$,

$$
P(X_{ni}=x)=
\frac{\exp\left\{\rho_{sg}\left[x\theta_n-
\sum_{k=1}^{x}\delta_{ik}\right]\right\}}
{\sum_{y=0}^{m_i}\exp\left\{\rho_{sg}\left[y\theta_n-
\sum_{k=1}^{y}\delta_{ik}\right]\right\}},
\qquad \rho_{sg}=\alpha_s\phi_g.
$$

Here $\alpha_s$ is the item-set unit and $\phi_g$ is the person-group unit.
A frame-defining factor sets the unit and cannot also be tested as a separate
DIF factor on that frame's virtual items.

The model should be used when a frame-dependent unit is part of the
substantive measurement account, not simply because an equal-unit model fits
poorly.

```{r simulate}
d <- simulate_efrm(n_per_group = 300, items_per_set = 8,
                   set_unit_ratio = 1.30, group_unit_ratio = 1.10,
                   seed = 12)
truth <- attr(d, "truth")
```

## Fit and inspect the links

```{r fit}
fit <- rasch_efrm(
  d,
  item_sets = truth$item_sets,
  groups = "group",
  id = "id",
  boot_reps = 60
)
fit
```

`phi_table` reports person-group units, `alpha_table` reports item-set units,
and `set_table` reports the linked set locations. Units are positive and are
most naturally compared as ratios. The linking tables and notes should be read
before interpreting common-unit item or person estimates.

```{r tables}
fit$phi_table
fit$alpha_table
fit$set_table
fit$linking
```

Group units are estimated from within-frame pairwise conditional calibrations.
Set units and locations require persons common to the linked sets. A
disconnected design cannot establish a common unit.

The set-linking step uses an error-corrected method-of-moments estimator based
on the true-score variance-ratio argument in Humphry (2005). It is distinct
from the likelihood proposed in section 5.3 of that thesis. The crossed,
multigroup, and polytomous forms are extensions implemented in this package.

## Uncertainty and comparison with equal units

The default hybrid standard errors combine the pairwise sandwich covariance,
a person bootstrap for the set-linking stage, and delta-method propagation.
The moments-based link carries a small finite-sample bias at designs of
this size -- recovered set units typically sit a few per cent above the
planted values -- which vanishes with more linking information.
The example uses 60 replicates to keep the vignette quick; a final analysis
should use enough replicates to stabilise its reported standard errors. Set
`se_method = "bootstrap"` to refit the complete pipeline on each person
resample.

```{r comparison}
fit$frames
fit$efrm_vs_rasch
```

The raw equal-unit composite-likelihood difference is descriptive. The omnibus
Wald tests in `efrm_vs_rasch$unit_omnibus` test the unit families; individual
unit contrasts are Holm-adjusted follow-ups.

## References

Andrich, D., and Marais, I. (2019). *A Course in Rasch Measurement Theory:
Measuring in the Educational, Social and Health Sciences*. Springer.

Humphry, S. M. (2005). *Maintaining a Common Arbitrary Unit in Social
Measurement*. PhD thesis, Murdoch University.

Humphry, S. M., and Andrich, D. (2008). Understanding the unit in the Rasch
model. *Journal of Applied Measurement*, 9(3), 249--264.

Rasch, G. (1960). *Probabilistic Models for Some Intelligence and Attainment
Tests*. Copenhagen: Danish Institute for Educational Research. (Expanded
edition, 1980, Chicago: University of Chicago Press.)

Rasch, G. (1961). On general laws and the meaning of measurement in
psychology. In *Proceedings of the Fourth Berkeley Symposium on Mathematical
Statistics and Probability* (Vol. 4, pp. 321--333). Berkeley: University of
California Press.
