---
title: "Rasch analysis of paired comparisons"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rasch analysis of paired comparisons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.2)
options(digits = 4)
```

```{r library}
library(rasch)
```

## Fit the Bradley--Terry--Luce model

Paired-comparison data record two objects and an observed preference. In the
Bradley--Terry--Luce model, the log odds of choosing object A over object B are
their location difference. The comparison graph must connect all objects;
otherwise their relative locations are not identified.

```{r fit}
d <- simulate_btl(n_objects = 7, n_judges = 12,
                  reps_per_pair = 20, seed = 5)
fit <- btl(d, object_a = "object_a", object_b = "object_b",
           winner = "winner", judge = "judge")
fit
fit$objects
```

Judge residuals describe agreement with the common object scale; they are not
person measures. Object fit, judge fit, targeting, and comparison information
address different parts of the design and should be considered together.

```{r judges}
fit$judges
judge_surprise(fit, "J1")
btl_information(fit)
```

## Check transitivity and residual structure

Circular triads identify local contradictions in the observed ordering.
Residual dimensionality asks whether comparisons contain a structured second
attribute after the primary scale is fitted.

```{r structure}
tr <- btl_transitivity(fit)
tr
dimensions <- btl_dimensionality(fit, reps = 20)
dimensions
```

```{r plots, fig.alt = "Paired-comparison transitivity and residual dimensionality diagnostics."}
par(mfrow = c(1, 2))
plot_btl_transitivity(tr)
plot_btl_scree(dimensions)
```

The simulation reference for dimensionality uses twenty replicates here to keep
the vignette quick. A final analysis should use enough replicates to stabilise
the reference distribution.

## Equate panels through common objects

`btl_equate` aligns two calibrations that share at least three objects. For two
fitted calibrations, drift inference is withheld until independent judges and
comparisons are stated explicitly.

```{r equating-example, eval = FALSE}
eq <- btl_equate(current_panel, reference_panel, independent = TRUE)
eq$table
eq$equated                 # reference panel on the current panel's origin
```

A bank table may be used in place of `reference_panel`. Marginal object
standard errors are not enough for drift tests because they omit the
covariance created by the bank's fitted origin. Attach the joint matrix as
`attr(bank, "cov_location")`, ordered like the bank rows or named by object;
otherwise the alignment is descriptive. A bank treated as fixed may instead
carry zero standard errors. Dependent panels require a joint or paired
bootstrap outside this function.

## Linked frames for paired comparisons

`btl_efrm` extends the paired-comparison model when judges belong to panels and
objects belong to linked sets whose units or origins may differ. Cross-set
comparisons establish the common scale. Without enough cross-set links, set
units and origins are not identified.

```{r efrm}
de <- simulate_btl_efrm(
  n_objects_per_set = 5, n_sets = 2,
  n_judges_per_panel = 6, n_panels = 2,
  reps_within = 15, reps_cross = 15,
  set_units = c(1, 1.3), set_origins = c(0, 0.6), seed = 9
)
ef <- btl_efrm(
  de, "object_a", "object_b", winner = "winner", judge = "judge",
  panels = "panel", object_sets = attr(de, "truth")$object_sets,
  se_method = "conditional"
)
ef$phi_table
ef$alpha_table
ef$kappa_table
```

For final inference, the default judge bootstrap resamples judges within
panels and refits both stages. It therefore propagates stage-one uncertainty
and retains dependence among a judge's comparisons. The parametric bootstrap
(`se_method = "bootstrap"`) instead draws independent outcomes from the fitted
model and is useful as a model-based sensitivity analysis. The conditional
option used above is a fast inspection method and does not propagate
stage-one uncertainty into the set-linking parameters. Read the omnibus unit
tests before the Holm-adjusted individual unit contrasts.
