---
title: "Rasch analysis of paired comparisons"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rasch analysis of paired comparisons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.2)
options(digits = 4)
```

```{r library}
library(rasch)
```

## Fit the Bradley--Terry--Luce model

Paired-comparison data record two objects and an observed preference. In the
Bradley--Terry--Luce model (Bradley and Terry 1952; Luce 1959), the log odds
of choosing object A over object B are their location difference:

$$
P(A\succ B)=\frac{\exp(\beta_A)}
{\exp(\beta_A)+\exp(\beta_B)}.
$$

This is the conditional form of the dichotomous Rasch model (Rasch 1960;
Andrich 1978). The comparison graph must connect all objects;
otherwise their relative locations are not identified.

For ordered comparisons, `btl` fits the adjacent-category extension

$$
\log\frac{P(Y=r)}{P(Y=r-1)}=\beta_A-\beta_B-\tau_r,
$$

with thresholds symmetric under reversal of presentation order (Tutz 1986).

```{r fit}
d <- simulate_btl(n_objects = 7, n_judges = 12,
                  reps_per_pair = 20, seed = 5)
fit <- btl(d, object_a = "object_a", object_b = "object_b",
           winner = "winner", judge = "judge")
fit
fit$objects
```

Judge residuals describe agreement with the common object scale; they are not
person measures. Object fit, judge fit, targeting, and comparison information
address different parts of the design and should be considered together.

```{r judges}
fit$judges
judge_surprise(fit, "J1")
btl_information(fit)
```

## Check transitivity and residual structure

Circular triads (Kendall and Babington Smith 1940) identify local
contradictions in the observed ordering.
Residual dimensionality asks whether comparisons contain a structured second
attribute after the primary scale is fitted.

```{r structure}
tr <- btl_transitivity(fit)
tr
dimensions <- btl_dimensionality(fit, reps = 20)
dimensions
```

```{r plots, fig.alt = "Paired-comparison transitivity and residual dimensionality diagnostics."}
par(mfrow = c(1, 2))
plot_btl_transitivity(tr)
plot_btl_scree(dimensions)
```

The simulation reference for dimensionality uses twenty replicates here to keep
the vignette quick. A final analysis should use enough replicates to stabilise
the reference distribution.

## Examine DIF across judge groups

`btl_dif` tests whether object locations differ across nominated judge factors.
The omnibus analysis uses judges as the independent units. A significant term
is resolved into factor-specific object locations and pairwise logit
differences. HC3 covariance allows the precision of judge means to vary with
their comparison workloads. Omnibus and pairwise inference require at least
eight judges and eight effective judges in each factor level; estimates remain
descriptive below that boundary. The tables report both counts.

```{r btl-dif-example, eval = FALSE}
judge_group <- setNames(panel_data$discipline, panel_data$judge)
bd <- btl_dif(fit, judge_group)
bd$summary
bd$sizes
```

## Equate panels through common objects

`btl_equate` aligns two calibrations that share at least three objects. For two
fitted calibrations, drift inference is withheld until independent judges and
comparisons are stated explicitly.

```{r equating-example, eval = FALSE}
eq <- btl_equate(current_panel, reference_panel, independent = TRUE)
eq$table
eq$equated                 # reference panel on the current panel's origin
```

A bank table may be used in place of `reference_panel`. Marginal object
standard errors are not enough for drift tests because they omit the
covariance created by the bank's fitted origin. Attach the joint matrix as
`attr(bank, "cov_location")`, ordered like the bank rows or named by object;
otherwise the alignment is descriptive. A bank treated as fixed may instead
carry zero standard errors. Dependent panels require a joint or paired
bootstrap outside this function.

## Linked frames for paired comparisons

`btl_efrm` combines the comparative judgement model with Humphry's extended
frame of reference structure (Humphry and Andrich 2008). Judges belong to
panels, and objects belong to linked sets. For object $k$ in set $s$,

$$
v_k=\alpha_s\beta_k+\kappa_s.
$$

A same-set comparison in panel $g$ has logit
$\phi_g(\beta_A-\beta_B)$; a cross-set comparison has logit
$\phi_g(v_A-v_B)$. Cross-set comparisons identify the set units and origins.
The cross-set likelihood holds the within-set locations and panel units fixed.
It estimates the set transformations directly from the comparison outcomes
and does not use the finite-grid person-distribution link in `rasch_efrm()`.

```{r efrm}
de <- simulate_btl_efrm(
  n_objects_per_set = 5, n_sets = 2,
  n_judges_per_panel = 6, n_panels = 2,
  reps_within = 15, reps_cross = 15,
  set_units = c(1, 1.3), set_origins = c(0, 0.6), seed = 9
)
ef <- btl_efrm(
  de, "object_a", "object_b", winner = "winner", judge = "judge",
  panels = "panel", object_sets = attr(de, "truth")$object_sets,
  se_method = "conditional"
)
ef$phi_table
ef$alpha_table
ef$kappa_table
```

The default judge bootstrap resamples judges within panels and refits both
stages. The parametric bootstrap (`se_method = "bootstrap"`) draws independent
outcomes from the fitted model. The conditional option used above reports
estimates and conditional standard errors but withholds probabilities because
it does not propagate stage-one uncertainty into the set link. With either
bootstrap, omnibus tests cover the unit families and individual contrasts are
Holm-adjusted follow-ups. Judge-bootstrap tests require at least six judges
and 5.5 effective judges in every panel, and eight of each on a set link.
Judge resamples are distributed over four workers by default, or fewer where
the system imposes a lower limit. Set `seed` to reproduce the resamples. The
parametric bootstrap remains serial because its refits are inexpensive. In the
application, frame estimation runs in the background and may be cancelled.

With 12 judges and 20 repetitions per pair, null rejection for the three unit
families was 3.3--5.3 per cent under the judge bootstrap and 3.0--6.7 per cent
under the independent-outcome bootstrap. The staged set-unit estimate has
small finite-sample attenuation when the within-set locations are imprecise:
log-unit bias was -0.041 at 20 repetitions per pair, -0.016 at 50 and -0.007
at 100. The bootstrap intervals retained nominal coverage at the 20-repetition
design.

## A worked analysis on real data

The `party_blocs_crisis.R` case study fits the comparative judgement frame
model to real paired comparisons between political parties, with judge
panels and ideological blocs as frames:

```{r casestudy, eval = FALSE}
file.edit(system.file("casestudies", "party_blocs_crisis.R",
                      package = "rasch"))
```

## References

Andrich, D. (1978). Relationships between the Thurstone and Rasch approaches
to item scaling. *Applied Psychological Measurement*, 2, 451--462.

Bradley, R. A., and Terry, M. E. (1952). Rank analysis of incomplete block
designs: I. The method of paired comparisons. *Biometrika*, 39, 324--345.

Humphry, S. M., and Andrich, D. (2008). Understanding the unit in the
Rasch model. *Journal of Applied Measurement*, 9(3), 249--264.

Kendall, M. G., and Babington Smith, B. (1940). On the method of paired
comparisons. *Biometrika*, 31(3/4), 324--345.

Tutz, G. (1986). Bradley-Terry-Luce models with an ordered response.
*Journal of Mathematical Psychology*, 30(3), 306--316.

Luce, R. D. (1959). *Individual Choice Behavior: A Theoretical Analysis*.
Wiley.

Rasch, G. (1960). *Probabilistic Models for Some Intelligence and Attainment
Tests*. Copenhagen: Danish Institute for Educational Research. (Expanded
edition, 1980, Chicago: University of Chicago Press.)
