---
title: "Simulation-based checks of Rasch diagnostics"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation-based checks of Rasch diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.2, dpi = 96
)
options(digits = 4)
```

```{r library}
library(rasch)
```

## 1. Why simulate

Simulation provides a direct way to examine whether an estimator recovers
known parameters and whether a diagnostic responds to the departure it is
intended to detect. The package therefore includes simulators for the main
model families of Rasch Measurement Theory (Rasch 1960): `simulate_rasch`,
`simulate_btl`, `simulate_mfrm`, and `simulate_efrm`. Each can generate model-conforming data or introduce a known
departure from the model.

Every simulator returns data ready for its fitting function, with the
generating parameters and any specified departures carried on the object as
`attr(x, "truth")`. The print method reports what was planted, so a dataset
never becomes anonymous once it leaves the call that made it.

```{r whatplanted}
d <- simulate_rasch(n_persons = 400, n_items = 10, seed = 101)
d                                   # the print method reports the plant
names(attr(d, "truth"))            # the truth travels with the data
```

With no departures requested, the data conform to the model. The remaining
sections introduce one departure at a time and examine the corresponding
diagnostic. The diagnostic conventions follow Andrich and Marais (2019).

## 2. A model-conforming baseline

It is useful to begin with parameter recovery when the model holds, so that
later departures can be interpreted against a model-conforming reference.
`sim_recovery` compares the parameters a fit recovers against the ones the
simulator planted (mean-centring the locations, which the model identifies
only up to an origin), and `plot_recovery` draws each true-versus-estimated
panel with its correlation and root-mean-square error.

```{r baseline}
fit <- rasch(d, id = "id")
rec <- sim_recovery(fit, d)
rec
```

```{r baseline-plot, fig.alt = "Recovery scatter plots of planted against recovered item difficulty and person ability."}
plot_recovery(rec)
```

The two panels tell different stories, and the difference is not a defect of
the estimator but the arithmetic of information. Each item difficulty pools
evidence from all 400 persons, so it is recovered tightly: the correlation is
near unity and the error is around a tenth of a logit. Each person ability, by
contrast, rests on only the ten items that person answered, so its weighted
likelihood estimate (Warm 1989) carries a standard error of roughly eight
tenths of a logit; the recovery correlation is correspondingly lower and the
root-mean-square error much larger. (The bias column reads NA by design:
locations are identified only up to the scale origin, so a mean difference
from the planted values is not an estimable quantity.) A short
test measures persons imprecisely *even when the model is exactly true* --- a
point worth holding onto before reading any person-level result from a
ten-item scale.

## 3. Planting item misfit

The `discrimination` argument scales an item's slope. A value above one makes
an item over-discriminate --- its responses become more deterministic than the
model expects, a Guttman-like pattern --- while a value below one makes it
under-discriminate, adding noise. The fit residuals should separate both from
the well-behaved items. Here one central item is given a slope of 2.5 and
another a slope of 0.4.

```{r discrim}
disc <- rep(1, 10); disc[5] <- 2.5; disc[6] <- 0.4
d2  <- simulate_rasch(400, 10, discrimination = disc, seed = 21)
fit2 <- rasch(d2, id = "id")
fit2$items[, c("item", "location", "infit_ms", "outfit_ms")]
```

The two planted items depart from the pack in opposite directions, as the
log-of-mean-square fit statistics (Andrich and Marais 2019, ch. 23) require.
The over-discriminating item (I05) produces standardised residuals that are
*too small* --- responses more predictable than the model allows --- so its
infit and outfit mean squares fall below one. The under-discriminating item
(I06) produces residuals that are too large, so its mean squares rise above
one. The remaining items sit near the expected value of one. With a
400-by-10 draw these mean squares are themselves estimates; what matters is
the direction of departure and that the planted items separate from the rest.

## 4. Planting DIF

Differential item functioning is planted with `dif`, naming the affected items
and the size of the group shift, with `n_groups = 2` so there is a group to
differ. Here item I06 is given a uniform shift of one logit for the second
group; the other nine items are invariant. `dif_anova` analyses the residuals
by group over automatically sized trait class intervals (Hagquist and Andrich
2017), and should flag I06 alone.

```{r dif}
d3   <- simulate_rasch(500, 10, dif = list(items = "I06", uniform = 1),
                       n_groups = 2, seed = 303)
fit3 <- rasch(d3, id = "id", factors = "group")
da   <- dif_anova(fit3)
da$summary[, c("item", "term", "F_uniform", "p_uniform_adj", "uniform_DIF")]
```

Only I06 carries a significant group term after the Benjamini-Hochberg
adjustment across items; the other nine are quiet, as invariant items should
be. The analysis of variance answers *whether* there is DIF. To ask *how much*,
and whether it matters on the measurement scale, `dif_size` resolves the item
into one copy per group, refits, and reports the distance between the resolved
locations in logits (Andrich and Marais 2019, ch. 16).

```{r difsize}
dif_size(fit3, "I06", by = "group")
```

The resolved gap recovers the planted shift of one logit and clears the
half-logit practical criterion. Flagging and magnitude are complementary: a
significant `dif_anova` term says the item is not invariant; the `dif_size`
logit difference says whether the non-invariance is large enough to act on.

## 5. Planting local dependence

Response dependence --- one item's answer partly following another's --- is
planted with `dependence`, naming the item pairs and a strength. It should
raise the residual correlation of the planted pair above the small negative
value expected under local independence. `residual_correlations` returns Yen's
(1984) Q3 and, following Christensen, Makransky and Horton (2017), the adjusted
Q3\* (each Q3 less the average off-diagonal value). There is no universal
critical value for adjusted Q3, so binary flagging is available only when the
analyst supplies a screening threshold and remains heuristic.

```{r dependence}
d4   <- simulate_rasch(500, 10,
                       dependence = list(pairs = list(c("I04", "I05")),
                                         strength = 1.8), seed = 41)
fit4 <- rasch(d4, id = "id")
rc   <- residual_correlations(fit4, flag = 0.20)
rc$average                          # near -1/(L-1) under independence
head(rc$pairs, 3)                   # the planted pair leads the table
rc$flagged
```

The planted I04--I05 pair sits at the top of the sorted table and clears the
chosen heuristic screen; the average off-diagonal value is close to the
$-1/(L-1)$ expected when the items are locally independent. As with DIF,
screening is one thing and magnitude another. `dependence_magnitude` puts the effect on the logit scale by
the resolution method of Andrich and Kreiner (2010): the dependent item is
split by the category of the item it follows, refitted, and the threshold
displacement read off.

```{r depmag}
dependence_magnitude(fit4, dependent = "I05", independent = "I04")
```

The estimated dependence is around three quarters of a logit and highly
significant --- a substantial displacement of I05's threshold by the response
to I04, not a marginal correlation. (The residual principal components of the
same fit, via `plot_scree`, are the complementary tool for a *second
dimension* rather than a dependent pair.)

## 6. Paired comparisons

The paired-comparison simulator plants faults of its own. `erratic_judges` sets
a proportion of judges who choose at random; their disorder should surface in
the judge fit residuals and in the transitivity of their choices. Here two of
eight judges are erratic.

```{r btl}
b  <- simulate_btl(n_objects = 7, n_judges = 8, reps_per_pair = 30,
                   erratic_judges = 0.25, seed = 61)
bt <- btl(b, "object_a", "object_b", winner = "winner", judge = "judge")
bt$judges[order(-bt$judges$fit_resid), ]
```

The two erratic judges (J1 and J2) carry the largest infit and outfit mean
squares and the largest positive fit residuals; the well-behaved judges sit
below one. The judge fit residual is the paired-comparison counterpart of
person fit, and here it is the sharpest instrument.

`btl_transitivity` asks the single-dimension question in its native
paired-comparison form. A Bradley-Terry-Luce scale implies that preferences
stack into one consistent order: if A beats B and B beats C then A should beat
C. A *circular triad* (A beats B, B beats C, C beats A) is a local
contradiction, and their rate is compared with the one quarter expected from
pure guessing (Kendall and Babington Smith 1940).

```{r transitivity}
tr <- btl_transitivity(bt)
tr$summary[, c("n_objects", "n_triples", "n_circular",
               "circular_rate", "consistency")]
tr$judges[, c("judge", "n_triples", "circular_rate", "consistency")]
```

Pooled across all judges the object scale is nearly transitive --- majority
verdicts stack into essentially one order --- so the overall consistency is
high. The per-judge index separates the individuals, with the erratic judges
falling toward the bottom of the table. With only a handful of complete triples
per judge this index is noisier than the fit residual, but the two agree in
direction. Where the concern is not judge noise but a second attribute steering
some contests, `btl_dimensionality` provides the residual-"swirl" analogue of
residual principal components.

## 7. A small Monte Carlo power estimate

The plant-and-detect loop scales up. `sim_replicate` calls a simulator many
times with successive seeds, returning the datasets as a list, so a diagnostic
can be run across all of them and its behaviour summarised. Asking how often a
planted departure is flagged turns that into a detection-power estimate. Here a
uniform DIF of 0.8 logits is planted on one item across ten datasets, and each
is fitted and passed to `dif_anova`.

```{r power}
batch <- sim_replicate(simulate_rasch, 10, n_persons = 400, n_items = 8,
                       dif = list(items = "I04", uniform = 0.8),
                       n_groups = 2, seed = 700)
flagged <- vapply(batch, function(dd) {
  s <- dif_anova(rasch(dd, id = "id", factors = "group"))$summary
  isTRUE(s$uniform_DIF[s$item == "I04"])
}, logical(1))
mean(flagged)                       # proportion of runs that flagged I04
```

The estimate is the proportion of the ten runs in which the planted item was
flagged after adjustment. A shift of 0.8 logits split over two groups of 200 is
detected in only about half of these runs: a real departure of moderate size
against a modest sample is caught only some of the time. Ten
replicates suffice to demonstrate the loop, not to pin the number down --- the
Monte Carlo error on ten draws is large --- but the same six lines, run with a
few hundred replicates and swept over sample size and effect, are a complete
power study.

## 8. Validation studies

The simulators above are also the package's own test bench. Each release is
validated by a simulation battery that runs every model family --- the
dichotomous model, the polytomous models, many-facet models, the extended
frame of reference model, and the comparative judgement models --- against
its full set of diagnostics, under complete data, 25% random missingness,
and linked structural designs: 234 recorded checks, of which 231 passed.
(The three failures are retained in the result table deliberately: two
exposed the standard-error defect corrected below, and one was a
mis-designed power scenario in the battery itself, redesigned and passing.)
Parameter recovery is
checked against the planted values; reported standard errors against the
empirical sampling variability of the estimates (all calibration ratios fall
within 0.85--1.20, most within a few percent of 1); significance tests
against their nominal rates under model-true data; and every diagnostic
against its planted departure for power. The identification guards are
exercised in both directions: connected linked designs fit and recover well,
while genuinely disconnected designs are refused rather than fitted. The
battery's scripts (seeds inline) and its per-check result table are kept in
the package sources under `tools/simval/`, so every number is reproducible
from the repository.

A second round followed the external review, targeting the newer
inferential procedures with the reporting standard of Morris, White and
Crowther (2019): every rate with its Monte Carlo standard error, replicate
accounting that separates attempted, refused, and non-converged fits, and
at least 1,000 replicates behind each principal null claim wherever a
replicate is computationally feasible (the tailored bootstrap, whose
replicates each run a complete bootstrap, is the stated exception; a few
follow-up tables record refusals and non-convergences as one combined
count and say so in their notes). Seven studies
(committed under `tools/simval/studies/`, results and provenance under
`tools/simval/results/`) covered the custom Wald and contrast tests,
judge-clustered comparative judgement inference, the tailored-analysis
bootstrap, the `lr_test` composite-likelihood comparison, equating
multiplicity, structural missingness, and person-measure coverage. Most of
it confirmed calibration: `lr_test` rejects a true rating-scale model at
4.65% over 2,000 replicates; equating's familywise drift error is nominal
at every anchor count tried (4.9--5.5% over 2,000--4,000 replicates), with
shift coverage 94.6--94.7% across bank modes; person-measure (WLE)
coverage sits at 0.945--0.983 across the central ability range for the
principal ten-item design (the twenty-five-item test dips to 0.933 at one
central point), and in the tails the intervals turn conservative rather
than failing (at true
abilities two logits past the item bank, bias approaches the distance to
the bank but coverage stays at or above nominal); balanced
judge-clustered designs are calibrated from ten judges up (Type I 5.01%,
coverage 94.5% at exactly ten judges, 1,200 replicates); and the
tailored-analysis bootstrap did not overstate at the design studied: its
familywise error is 0.8% over 240 full-procedure replicates (anchor
selection re-run inside every one -- fewer than the battery's usual
thousand, because each replicate runs a complete bootstrap and costs
minutes), with low power against planted guessing at feasible bootstrap
sizes, which is why tailored shifts are descriptive by default and the
bootstrap warns about its own resolution floor. Its power grid (four to
six replicates per cell) bounds behaviour rather than estimating it.

The second round also caught three defects, each confirmed by an
adversarial verification pass, diagnosed, corrected, and re-validated:

- The extended frame of reference model's set-unit tests rejected a true
  null at 9.4% (1,000 replicates): the linking-stage bootstrap resampled
  persons but held the estimated thresholds fixed, and the set unit is a
  scale, so calibration noise moves every person estimate's variance
  coherently. Each bootstrap replicate now redraws the thresholds and
  panel units jointly from their estimated stage-1 covariance
  (cross-covariance preserved); the corrected tests reject at 4.9%
  (1,200 fresh replicates), stable across item counts, sample sizes,
  imbalance, a third linked set, and weak linking.
- Judge-clustered comparative judgement inference was calibrated for the
  judge counts tested but not for concentrated allocations: one judge
  performing half the comparisons leaves about four effective clusters
  whatever the nominal count, and the true-null rejection climbs to ~9%.
  The guard now withholds clustered inference below eight effective
  judges (inverse Simpson of the comparison shares), cautions between
  eight and nine and a half, and reports `cl$n_units_effective`.
- The paired-comparison extended-frame unit tests used chi-square and
  normal references on covariances estimated from ten to twenty judges;
  the set-origin omnibus rejected at 8.7%. Judge-limited references
  (Hotelling-style F for the omnibus, t for the unit tables) restore
  5.5%.

Two suspected defects were exonerated by the same machinery: the
many-facet interaction omnibus at 25 parameters is calibrated (4.3% and
5.2% at 600 fixed-truth replicates), and a reported two-fold standard
error understatement for partial credit thresholds under a rare category
proved to be mostly an artifact of pooling heterogeneous items -- though
the per-item decomposition it prompted exposed one genuine case, a
four-fold understatement in a threshold neighbouring a ~4-response
category, and the weak-category guard now withholds a whole item's
standard errors when any category falls below eight responses (the
survivors it still reports are calibrated to 1.01--1.08).

One correction came out of the first battery. `dependence_magnitude` followed
Andrich and Marais (2019, eqs. 24.9--24.11) in pooling the two resolved
items' variances as if their estimates were independent, on the grounds that
disjoint persons answer them. In the joint refit, however, the two estimates
are negatively correlated through the shared comparator items, and the
pooled standard error is too small: with ten items, the null rejection rate
was 7.5% at a nominal 5% over 1,200 replicates. Since release 1.14.2 the
standard error is the delta-method contrast over the fit's sandwich
covariance, which restores the nominal rate: 4.8% over 400 fresh-seed
replicates (Monte Carlo error about 1 percentage point), 3.5% for the
partial credit case, and 5.4% when the corrected formula is applied to the
original 1,200 diagnostic replicates. The estimate of *d* itself is
unaffected.

## References

Andrich, D., and Kreiner, S. (2010). Quantifying response dependence between
two dichotomous items using the Rasch model. *Applied Psychological
Measurement*, 34, 181--192.

Andrich, D., and Marais, I. (2019). *A Course in Rasch Measurement Theory*.
Springer.

Christensen, K. B., Makransky, G., and Horton, M. (2017). Critical values for
Yen's Q3. *Applied Psychological Measurement*, 41, 178--194.

Hagquist, C., and Andrich, D. (2017). Recent advances in analysis of
differential item functioning in health research using the Rasch model.
*Health and Quality of Life Outcomes*, 15, 181.

Kendall, M. G., and Babington Smith, B. (1940). On the method of paired
comparisons. *Biometrika*, 31, 324--345.

Morris, T. P., White, I. R., and Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine*, 38,
2074--2102.

Rasch, G. (1960). *Probabilistic Models for Some Intelligence and Attainment
Tests*. Copenhagen: Danish Institute for Educational Research. (Expanded
edition, 1980, Chicago: University of Chicago Press.)

Warm, T. A. (1989). Weighted likelihood estimation of ability in item response
theory. *Psychometrika*, 54, 427--450.

Yen, W. M. (1984). Effects of local item dependence on the fit and equating
performance of the three-parameter logistic model. *Applied Psychological
Measurement*, 8, 125--145.
