---
title: "Simulation-based checks of Rasch diagnostics"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation-based checks of Rasch diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.2, dpi = 96
)
options(digits = 4)
```

```{r library}
library(rasch)
```

## Simulating from the models

Simulation is useful when the sampling behaviour of an estimate or diagnostic
depends on the test design. The package includes simulators for the ordinary
Rasch models (Rasch 1960; Andrich and Marais 2019), many-facet models,
extended frames of reference (Humphry 2005; Humphry and Andrich 2008), and
comparative judgement of paired
comparisons. They can generate model-conforming data or introduce a specified
departure.

Each simulator stores the generating values in `attr(x, "truth")`.

```{r baseline-data}
d <- simulate_rasch(n_persons = 400, n_items = 10, seed = 101)
names(attr(d, "truth"))
```

## Parameter recovery

`sim_recovery` compares fitted parameters with the generating values. Location
parameters are centred before comparison because their origin is arbitrary.

```{r baseline}
fit <- rasch(d, id = "id")
rec <- sim_recovery(fit, d)
rec
```

```{r baseline-plot, fig.alt = "Planted and recovered item and person locations."}
plot_recovery(rec)
```

Item locations pool information over persons. Each person location is based on
the items answered by that person, so person recovery is usually less precise
on a short test. This difference should be judged against the reported
standard errors rather than the raw recovery correlations alone.

## Item misfit

The `discrimination` argument changes an item's response slope. Values above
one produce more deterministic responses than the Rasch model expects; values
below one produce less deterministic responses.

```{r discrim}
disc <- rep(1, 10)
disc[5] <- 2.5
disc[6] <- 0.4

d2 <- simulate_rasch(400, 10, discrimination = disc, seed = 21)
fit2 <- rasch(d2, id = "id")
fit2$items[, c("item", "location", "infit_ms", "outfit_ms")]
```

Over-discrimination tends to give mean-square statistics below one;
under-discrimination tends to give values above one. Their sampling variation
still depends on the item location, sample, and test length.

## Differential item functioning

The `dif` argument shifts selected items for a person group. Here I06 differs
by one logit in the second group.

```{r dif}
d3 <- simulate_rasch(
  500, 10,
  dif = list(items = "I06", uniform = 1),
  n_groups = 2,
  seed = 303
)

fit3 <- rasch(d3, id = "id", factors = "group")
da <- dif_anova(fit3)
da$summary[, c("item", "term", "F_uniform", "p_uniform_adj", "uniform_DIF")]
```

`dif_anova` tests invariance. `dif_size` resolves the item by group and reports
the difference between the resolved locations in logits.

```{r dif-size}
dif_size(fit3, "I06", by = "group")
```

A simulation study of DIF should record false-positive rates for invariant
items as well as detection of the shifted item. Sample size, group imbalance,
targeting, test length, and shift size should be varied separately.

## Local response dependence

The `dependence` argument makes one item's response partly follow another.
`residual_correlations` reports Yen's Q3 and adjusted Q3. Because adjusted Q3
has no universal critical value, `flag` is a screening threshold supplied by
the analyst (Yen 1984; Christensen, Makransky and Horton 2017).

```{r dependence}
d4 <- simulate_rasch(
  500, 10,
  dependence = list(
    pairs = list(c("I04", "I05")),
    strength = 1.8
  ),
  seed = 41
)

fit4 <- rasch(d4, id = "id")
rc <- residual_correlations(fit4, flag = 0.20)
head(rc$pairs, 3)
```

`dependence_magnitude` resolves the dependent item by the response to the
independent item and expresses the displacement on the logit scale (Andrich
and Kreiner 2010).

```{r dependence-size}
dependence_magnitude(fit4, dependent = "I05", independent = "I04")
```

## Paired comparisons

The paired-comparison simulator can introduce erratic judges, ties, position
effects, or within-judge dependence. In this example, one quarter of the
judges respond at random.

```{r btl}
b <- simulate_btl(
  n_objects = 7,
  n_judges = 8,
  reps_per_pair = 30,
  erratic_judges = 0.25,
  seed = 61
)

bt <- btl(b, "object_a", "object_b", winner = "winner", judge = "judge")
bt$judges[order(-bt$judges$fit_resid), ]
```

Judge fit describes agreement with the common object scale. Transitivity is a
different summary: it counts circular triads in the observed comparisons.

```{r transitivity}
tr <- btl_transitivity(bt)
tr$summary
head(tr$judges)
```

## Repeated simulation

`sim_replicate` generates datasets with successive seeds. The same analysis
can then be applied to each dataset to estimate bias, coverage, rejection
rates, or power.

```{r power}
batch <- sim_replicate(
  simulate_rasch, 10,
  n_persons = 400,
  n_items = 8,
  dif = list(items = "I04", uniform = 0.8),
  n_groups = 2,
  seed = 700
)

flagged <- vapply(batch, function(dd) {
  s <- dif_anova(rasch(dd, id = "id", factors = "group"))$summary
  isTRUE(s$uniform_DIF[s$item == "I04"])
}, logical(1))

mean(flagged)
```

Ten replicates demonstrate the workflow but do not give a stable power
estimate. For a Monte Carlo proportion $\hat p$ based on $R$ independent
replicates, the estimated Monte Carlo standard error is

$$
\operatorname{MCSE}(\hat p)=
\sqrt{\frac{\hat p(1-\hat p)}{R}}.
$$

The number of attempted, refused, and non-converged fits should be reported.
Bias and coverage should be calculated for each generating condition rather
than after pooling conditions with different true values.

## Validation studies

The repository contains the simulation studies used to check parameter
recovery, standard errors, confidence-interval coverage, null rejection rates,
power, and identification guards. The scripts and result tables are under
`tools/simval/`; they are excluded from the CRAN source package for size,
and are computationally intensive to re-run. The examples in this vignette
use small runs and are intended as templates for design-specific studies.

The principal calibration results, each carried with its script and
provenance in the result tables:

| Quantity | Design | Result |
|---|---|---|
| `lr_test` adjusted size | 500 persons, 8 items, 3 categories | 4.7% at the 0.05 level (2,000 replicates) |
| `lr_test` small-sample edge | 300 persons, 12 items, 4 categories | 6.1% among 1,927/2,000 admissible replicates |
| `dependence_magnitude` size | 800 persons, 10 items | 7.5% pooled-variance (pre-fix) to 4.8% covariance-based |
| EFRM set-unit omnibus size | 200/group, 8 items/set, 2 sets | 9.4% (pre-fix) to 4.9% (1,200 replicates, eight designs) |
| Comparative judgement contrasts | 10-50 judges, balanced | 5.0% size, 94.5% coverage (1,200 replicates) |
| Effective-judge thresholds | one judge with 15-50% of comparisons | ~9% at 4 effective, ~7% at 6-7, nominal when balanced |
| Equating familywise error | 3, 5, and 10 anchors | 4.9-5.5% (2,000-4,000 replicates) |
| Person-measure coverage | 10-item test, central range | 0.945-0.983; conservative in the tails |
| Tailored bootstrap familywise error | 300 persons, 8 items | 0.8% over 240 full-procedure replicates |
| CL-AIC model selection | PCM vs RSM; free vs PC thresholds (items and CJ) | null false selection 4.5-5.2% multi-parameter, ~17% one-parameter (the theoretical AIC rates); detection 95-100% at strong departures |
| Paired-comparison effect tests | 8 objects, 14 judges | position/exposure nulls 5.8%/5.9%; carry-over 8.3% at 14 judges, 5.3% at 30; power 62/39/77% at 0.6 logits |
| Cross-package agreement | sirt, eRm, TAM, BradleyTerry2, VGAM, lme4 | identical-likelihood comparators at solver precision; estimator variants within 0.02-0.15 logits; unit estimators tracked against external anchors |
| EFRM unit linking (corrected) | 5-15 items/set; ratios 1-2; skew 0.5-2.8; booklet and pairwise-overlap designs | bias within ±0.015 throughout; null size 2.7-4.8%; coverage 93-99%; stays unbiased under skew where a normal-population anchor drifts |

## References

Andrich, D. (1978). Relationships between the Thurstone and Rasch approaches
to item scaling. *Applied Psychological Measurement*, 2(3), 451--462.

Andrich, D., and Kreiner, S. (2010). Quantifying response dependence between
two dichotomous items using the Rasch model. *Applied Psychological
Measurement*, 34, 181--192.

Andrich, D., and Marais, I. (2019). *A Course in Rasch Measurement Theory*.
Springer.

Christensen, K. B., Makransky, G., and Horton, M. (2017). Critical values for
Yen's Q3. *Applied Psychological Measurement*, 41, 178--194.

Morris, T. P., White, I. R., and Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine*, 38,
2074--2102.

Humphry, S. M. (2005). *Maintaining a Common Arbitrary Unit in Social
Measurement*. PhD thesis, Murdoch University.

Humphry, S. M., and Andrich, D. (2008). Understanding the unit in the
Rasch model. *Journal of Applied Measurement*, 9(3), 249--264.

Rasch, G. (1960). *Probabilistic Models for Some Intelligence and Attainment
Tests*. Danish Institute for Educational Research. Expanded edition,
University of Chicago Press, 1980.

Tutz, G. (1986). Bradley-Terry-Luce models with an ordered response.
*Journal of Mathematical Psychology*, 30(3), 306--316.

Warm, T. A. (1989). Weighted likelihood estimation of ability in item response
theory. *Psychometrika*, 54, 427--450.

Yen, W. M. (1984). Effects of local item dependence on the fit and equating
performance of the three-parameter logistic model. *Applied Psychological
Measurement*, 8, 125--145.
