---
title: "A Rasch analysis workflow"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Rasch analysis workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.2)
options(digits = 4)
```

```{r library}
library(rasch)
```

This vignette follows a Rasch analysis from the overall summary to item and
person fit, targeting, dependence, and differential item functioning (DIF).
The order matters. A fit statistic is difficult to interpret without knowing
whether the scale separates the sample or supplies information over the
relevant part of the latent trait.

The same analyses are available in the Shiny application. Run
`rasch::run_app()` to import data, fit a model, inspect the results, and export
tables or plots without writing R analysis code.

## Fit the model required by the scoring structure

`rasch` fits the partial credit model by default (Rasch 1960; Andrich and
Marais 2019). Dichotomous data are its one-threshold special case. Set
`model = "RSM"` only when the same category threshold structure is intended
to hold across items.

For person $n$, item $i$, and score $x=0,\ldots,m_i$, the partial credit model
is

$$
P(X_{ni}=x)=
\frac{\exp\left\{x\theta_n-\sum_{k=1}^{x}\delta_{ik}\right\}}
{\sum_{y=0}^{m_i}\exp\left\{y\theta_n-
\sum_{k=1}^{y}\delta_{ik}\right\}}.
$$

The rating scale model constrains $\delta_{ik}=\beta_i+\tau_k$. Pairwise
conditioning removes $\theta_n$ from the item likelihood. Person locations
are subsequently estimated by Warm's weighted likelihood method (Warm 1989).

The example is polytomous and contains three person groups. One item has
disordered generating thresholds, one has uniform DIF, and one item pair has
local response dependence. These departures make the diagnostic sequence
visible without changing the commands used for observed data. Before fitting
observed data, check the coding and frequency of every category. Negative
scores are read as missing; valid categories begin at zero.

```{r fit}
d <- simulate_rasch(
  n_persons = 600,
  n_items = 12,
  model = "PCM",
  n_categories = 4,
  difficulty = c(-1.5, 1.5),
  disordered = "I04",
  dependence = list(pairs = list(c("I10", "I11")), strength = 1.3),
  dif = list(items = "I08", uniform = 0.8),
  n_groups = 3,
  seed = 17
)

fit <- rasch(d, model = "PCM", id = "id", factors = "group")
```

## Overall summary and reliability

The summary establishes whether estimation converged and gives the overall
item--trait interaction, reliability, and the package's qualitative assessment
of the power of fit tests. It should be read before individual item or person
results.

```{r overall-summary}
fit
```

The person separation index (PSI) compares the observed variance of person
locations with their mean error variance:

$$
\mathrm{PSI}=
\frac{\operatorname{Var}(\hat\theta)-
\operatorname{mean}\{\operatorname{SE}(\hat\theta)^2\}}
{\operatorname{Var}(\hat\theta)}.
$$

The implementation truncates negative values at zero. `fit$psi_noext` removes
persons with extreme scores and is useful when extremes inflate the observed
spread. The separation ratio and number of strata express the same information
on more interpretable scales.

The power label is a screening judgement based on PSI, not a formal power
calculation. Low reliability weakens the ordering of persons and the formation
of distinct class intervals, so item--trait and DIF tests may fail to detect
real departures. Power also depends on sample size, test length, targeting,
category use, and trait spread. Conversely, a large sample can make a small
departure statistically significant. Here the PSI is `r sprintf("%.2f", fit$psi$PSI)`,
which the package classifies as `r fit$power_of_fit`; weak separation is
therefore not an immediate explanation for a quiet fit test. Fit residuals,
effect sizes, plots, and substantive importance remain necessary.

## Item estimates, fit, and thresholds

Item locations and their standard errors should be read alongside fit
residuals and the Holm-adjusted item--trait probabilities. The table shows the
six items with the largest absolute residuals; the complete results are in
`fit$items`. A positive residual indicates more variation than expected and a
negative residual indicates responses that are more predictable than expected.
The conventional $\pm2.5$ band is a screening rule rather than a separate
hypothesis test.

```{r item-estimates}
item_order <- order(abs(fit$items$fit_resid), decreasing = TRUE)
head(fit$items[item_order, c(
  "item", "location", "se", "fit_resid", "p_adj"
)], 6)
```

```{r item-fit-plot, fig.alt = "Item locations plotted against item fit residuals."}
plot_item_map(fit)
```

For a polytomous item, successive thresholds should normally increase on the
latent scale. The threshold map shows their order directly. Item I04 has an
intervening category without its own region on the scale; the category curves
show the same problem in probability form.

```{r thresholds, fig.height = 5.2, fig.alt = "Estimated category thresholds for all items on the common logit scale."}
plot_threshold_map(fit)
```

```{r category-curves, fig.alt = "Category characteristic curves and observed category proportions for item I04."}
plot_ccc(fit, "I04", observed = TRUE)
```

Disordering alone does not justify collapsing categories. The response-option
meanings, observed use, and category curves should support any rescoring. The
model must then be refitted and checked again.

## Person estimates and fit

Person fit addresses the consistency of each response pattern with the fitted
scale. The following table puts the largest absolute residuals first. Positive
residuals indicate unexpectedly erratic patterns; negative residuals indicate
patterns that are unusually predictable.

```{r person-estimates}
person_order <- order(abs(fit$person$fit_resid),
                      decreasing = TRUE, na.last = TRUE)
head(fit$person[person_order, c(
  "id", "group", "raw", "theta", "se", "fit_resid"
)], 6)
```

```{r person-fit-plot, fig.alt = "Person locations plotted against person fit residuals."}
plot_person_fit(fit)
```

An unexpected response pattern may reflect coding or data-entry errors,
careless responding, a secondary trait, or a genuine but unusual person. It is
not, by itself, a reason to remove the person. In this example,
`r sum(abs(fit$person$fit_resid) > 2.5, na.rm = TRUE)` persons
(`r sprintf("%.1f", 100 * mean(abs(fit$person$fit_resid) > 2.5, na.rm = TRUE))`%)
fall outside the displayed band; the plot reports the same count and
percentage. The `statistic` argument displays the standardised infit or
outfit in place of the fit residual, under the same band. Fit residuals are
unavailable for extreme response patterns because those patterns do not
provide an interior location at which fit can be assessed.

## Targeting and information

Targeting concerns the match between the person distribution and the item
threshold distribution. The table reports their locations and spread, the
proportions of persons beyond the threshold range, and the principal
reliability indices.

```{r targeting-summary}
fit$targeting
```

For a Rasch model, test information is the sum of the conditional response
variances:

$$
I(\theta)=\sum_i \operatorname{Var}(X_i\mid\theta),
\qquad
\operatorname{SE}(\hat\theta)\approx I(\theta)^{-1/2}.
$$

The person--item map below places person locations and item thresholds on the
same logit scale. The information curve shows where the test is most precise.

```{r targeting-map, fig.height = 5.2, fig.alt = "Person and item distributions with the test information curve on the common logit scale."}
plot_pimap(fit, information = TRUE)
```

The Wright map shows the same alignment in the conventional vertical
arrangement: the person distribution beside the item thresholds on one logit
scale.

```{r wright, fig.height = 5.2, fig.alt = "Wright map of the person distribution and item thresholds on the common logit scale."}
plot_wright(fit)
```

The optional `WrightMap` package draws the same map with greater flexibility,
including several person and item panels (Torres Irribarra and Freund 2025).
Polytomous thresholds are labelled `t1`, `t2`, and so on. Threshold labels are
omitted for a wholly dichotomous scale; a dichotomous item in a mixed scale
retains `t1`. Panels should answer a substantive question; here the person
distributions are separated by the fitted group factor.

```{r wrightmap, fig.width = 8, fig.height = 5.2, fig.alt = "Wright map with one person panel per group."}
if (requireNamespace("WrightMap", quietly = TRUE)) {
  wright_map(fit, person_panels = "group")
}
```

If `WrightMap` is not installed, install it with
`install.packages("WrightMap")` before running this chunk.

## Local and trait dependence

Local response dependence occurs when two responses remain associated after
conditioning on the latent trait. Yen's $Q3$ is the correlation between two
items' standardised residuals. Because raw $Q3$ values have a negative
baseline in a finite test, `q3_star` subtracts the average off-diagonal value.

```{r local-dependence}
q3 <- residual_correlations(fit)
q3$average
head(q3$pairs[, c("item_a", "item_b", "q3", "q3_star")], 5)
```

```{r local-dependence-plot, fig.alt = "Heatmap of adjusted residual correlations between items."}
plot_resid_cor(fit)
```

There is no universal critical value for adjusted $Q3$ (Christensen,
Makransky and Horton 2017). Its size, the response process, and the content of
the item pair matter. When theory supports treating dependent items as one
superitem, `combine_items()` refits the complete calibration and
`spread_test()` compares its threshold spread with the binomial bound. The
bound does not apply to a superitem containing a polytomous component.

Trait dependence is examined through the residual components and by comparing
person estimates from opposed item subsets (Smith 2002). For subsets $A$ and
$B$, the person-level statistic is

$$
t_n=\frac{\hat\theta_{nA}-\hat\theta_{nB}}
{\sqrt{\operatorname{SE}(\hat\theta_{nA})^2+
       \operatorname{SE}(\hat\theta_{nB})^2}}.
$$

Under unidimensionality, about the nominated alpha level of these comparisons
should be significant. The exact binomial interval, the residual loadings,
and the score points in each subset are part of the result.

```{r trait-dependence}
dimensionality <- dimensionality_test(fit)
```

```{r trait-dependence-plot, fig.alt = "Loadings of items on the first residual component."}
plot_pca(fit)
```

Here, `r sprintf("%.1f", 100 * dimensionality$prop_significant)`% of the
person comparisons are significant (exact 95% interval
`r sprintf("%.1f", 100 * dimensionality$ci[1])`% to
`r sprintf("%.1f", 100 * dimensionality$ci[2])`%). The test does not flag
trait dependence, but one opposed subset contains only
`r min(dimensionality$score_points)` score points. A quiet result from a short
subtest is inconclusive rather than evidence that a secondary trait is absent.

## Differential item functioning

Person factors must be nominated when the model is fitted. With a person
factor $G$ and trait class interval $C$, the residual model is

$$
z=\mu+G+C+G\mathbin{:}C+\varepsilon.
$$

The factor term tests uniform DIF; the interaction with class interval tests
non-uniform DIF. `dif_anova()` fits all nominated factors together, recognises
within-person factors in repeated designs, and applies Holm familywise
correction over the item-by-term tests.

```{r dif}
dif <- dif_anova(fit, sizes = TRUE)
flagged_dif <- subset(dif$summary, uniform_DIF | nonuniform_DIF)
flagged_dif[, c(
  "item", "term", "F_uniform", "p_uniform_adj", "eta2_uniform",
  "F_nonuniform", "p_nonuniform_adj", "eta2_nonuniform"
)]
```

```{r dif-plot, fig.alt = "Observed and expected item characteristic curves for item I08 by person group."}
plot_icc(fit, "I08", group = "group")
```

For a significant factor with more than two levels, the follow-up should
estimate the relevant differences in Rasch logits rather than rely on a
generic Tukey procedure. With `sizes = TRUE`, `dif_anova()` returns the
Holm-adjusted marginal pairwise comparisons for significant main effects and
difference-in-differences for significant interactions.

```{r dif-follow-up}
dif$posthoc[, c(
  "item", "contrast", "estimate", "se", "p_adj",
  "lower", "upper", "practical"
)]
```

Statistical significance and practical magnitude answer different questions.
Any split must be supported by the response process and should be applied with
`resolve_dif()`, which refits the calibration and updates the item and person
estimates. The revised fit then goes through the same summary, fit, targeting,
dependence, and DIF sequence.

## References

Andrich, D., and Marais, I. (2019). *A Course in Rasch Measurement Theory:
Measuring in the Educational, Social and Health Sciences*. Springer.

Christensen, K. B., Makransky, G., and Horton, M. (2017). Critical values for
Yen's Q3: Identification of local dependence in the Rasch model using residual
correlations. *Applied Psychological Measurement*, 41(3), 178--194.

Holm, S. (1979). A simple sequentially rejective multiple test procedure.
*Scandinavian Journal of Statistics*, 6(2), 65--70.

Rasch, G. (1960). *Probabilistic Models for Some Intelligence and Attainment
Tests*. Copenhagen: Danish Institute for Educational Research. (Expanded
edition, 1980, Chicago: University of Chicago Press.)

Smith, E. V. Jr. (2002). Detecting and evaluating the impact of
multidimensionality using item fit statistics and principal component analysis
of residuals. *Journal of Applied Measurement*, 3(2), 205--231.

Torres Irribarra, D., and Freund, R. (2025). *WrightMap: IRT item-person map
with ConQuest integration*. R package version 1.5.

Warm, T. A. (1989). Weighted likelihood estimation of ability in item
response theory. *Psychometrika*, 54(3), 427--450.

Yen, W. M. (1984). Effects of local item dependence on the fit and equating
performance of the three-parameter logistic model. *Applied Psychological
Measurement*, 8(2), 125--145.
