---
title: "A Rasch analysis workflow"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Rasch analysis workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.2)
options(digits = 4)
```

```{r library}
library(rasch)
```

This vignette walks through the standard analysis sequence of Rasch
Measurement Theory and its model suite (Rasch 1960, 1961; Andrich and
Marais 2019; for the frame of reference models Humphry and Andrich 2008,
and for comparative judgement as the conditional form of the dichotomous
Rasch model, Andrich 1978): fit by
pairwise conditional estimation, then examine whether the data meet the
model's requirements before the measures are interpreted.

The same principal workflow is available through the package's point-and-click
Shiny application. Run `rasch::run_app()` after installation to import data,
choose a model, inspect diagnostics, and export results without writing R
analysis code. The application shows the corresponding R call for each
analysis; this article explains the same decisions in their command-line form.

## Fit the model required by the scoring structure

`rasch` fits the partial credit model by default. Dichotomous data are its
one-threshold special case. Set `model = "RSM"` only when the same category
threshold structure is intended to hold across items.

The example uses simulated data so that the complete workflow is reproducible.
With observed data, inspect the coding and frequency of every category before
fitting the model. Negative scores are read as missing; valid categories begin
at zero.

```{r fit}
d <- simulate_rasch(n_persons = 400, n_items = 10, seed = 17)
fit <- rasch(d, id = "id")
fit
```

## Read fit and targeting together

Item fit is not a single pass-or-fail decision. Examine the direction and size
of residual misfit, the item-trait interaction, category thresholds, and the
substantive content of an item. Targeting determines how much information the
scale supplies over the observed person distribution.

```{r summaries}
fit_summary_table(fit)
targeting_table(fit)
fit$threshold_diagnostics
head(test_information(fit))
```

```{r targeting, fig.alt = "Person and item distributions on the common logit scale."}
plot_pimap(fit)
```

## Check invariance, independence, and dimensionality

Person factors should be nominated in the fit and tested jointly with
`dif_anova`. Residual correlations and the dimensionality test address local
independence and the adequacy of one measurement dimension. They are evidence
to investigate, not automatic instructions to delete or combine items.

```{r structure}
head(residual_correlations(fit))
dimensionality_test(fit)
```

Binary residual-correlation flags require an analyst-supplied heuristic
threshold because adjusted Q3 has no universal critical value. The
dimensionality test also withholds a categorical verdict unless each opposed
subtest contains at least 15 score points; this is preferable to treating a
short, low-information split as evidence for or against one dimension.

## Link calibrations through common items

Two separately fitted tests can be aligned through their common items. Drift
tests are meaningful only when the calibrations use independent sampling
units; state this rather than allowing the software to assume it.

```{r equating-example, eval = FALSE}
eq <- equate_tests(current_fit, reference_fit, independent = TRUE)
eq$table
```

A saved item bank needs more than marginal standard errors for drift
inference. Its joint item-location covariance carries the dependence created
by the fitted origin and is attached as `attr(bank, "cov_location")`, in bank
row order or named by item. Without that matrix, the package still estimates
the origin shift and reports the link descriptively. A bank intended to be
fixed may instead use zero standard errors. For polytomous items, include the
maximum score in a `max` column so that scoring-scale compatibility can be
checked.

## Report the final calibration

After any substantively defensible change, refit and repeat the diagnostic
sequence. A report should state the model, scoring and missing-data rules,
sample and item exclusions, targeting, item and person fit, threshold
functioning, dependence, dimensionality, DIF, reliability, and the uncertainty
of person estimates. The score table supplies the complete-response conversion
from raw score to logits.

```{r reporting}
score_table(fit)
fit$psi
```

## References

Andrich, D. (1978). Relationships between the Thurstone and Rasch
approaches to item scaling. *Applied Psychological Measurement*, 2(3),
449--460.

Andrich, D., and Marais, I. (2019). *A Course in Rasch Measurement Theory:
Measuring in the Educational, Social and Health Sciences*. Springer.

Andrich, D., and Luo, G. (2003). Conditional pairwise estimation in the
Rasch model for ordered response categories using principal components.
*Journal of Applied Measurement*, 4(3), 205--221.

Humphry, S. M., and Andrich, D. (2008). Understanding the unit in the
Rasch model. *Journal of Applied Measurement*, 9(3), 249--264.

Rasch, G. (1960). *Probabilistic Models for Some Intelligence and Attainment
Tests*. Copenhagen: Danish Institute for Educational Research. (Expanded
edition, 1980, Chicago: University of Chicago Press.)

Rasch, G. (1961). On general laws and the meaning of measurement in
psychology. In *Proceedings of the Fourth Berkeley Symposium on Mathematical
Statistics and Probability* (Vol. 4, pp. 321--333). Berkeley: University of
California Press.

Warm, T. A. (1989). Weighted likelihood estimation of ability in item
response theory. *Psychometrika*, 54(3), 427--450.
