---
title: "A Rasch analysis workflow"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Rasch analysis workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.2)
options(digits = 4)
```

```{r library}
library(rasch)
```

This vignette follows a Rasch analysis from the overall summary to item and
person fit, targeting, dependence, and differential item functioning (DIF).
The order matters. A fit statistic is difficult to interpret without knowing
whether the scale separates the sample or supplies information over the
relevant part of the latent trait.

The same analyses are available in the Shiny application. Run
`rasch::run_app()` to import data, fit a model, inspect the results, and export
tables or plots without writing R analysis code.

## Fit the model required by the scoring structure

`rasch` fits the partial credit model by default. Dichotomous data are its
one-threshold special case. Set `model = "RSM"` only when the same category
threshold structure is intended to hold across items.

For person $n$, item $i$, and score $x=0,\ldots,m_i$, the partial credit model
is

$$
P(X_{ni}=x)=
\frac{\exp\left\{x\theta_n-\sum_{k=1}^{x}\delta_{ik}\right\}}
{\sum_{y=0}^{m_i}\exp\left\{y\theta_n-
\sum_{k=1}^{y}\delta_{ik}\right\}}.
$$

The rating scale model constrains $\delta_{ik}=\beta_i+\tau_k$. Pairwise
conditioning removes $\theta_n$ from the item likelihood. Person locations
are subsequently estimated by Warm's weighted likelihood method (Warm 1989).

The example is polytomous and contains three person groups. One item has
disordered generating thresholds, one has uniform DIF, and one item pair has
local response dependence. These departures make the diagnostic sequence
visible without changing the commands used for observed data. Before fitting
observed data, check the coding and frequency of every category. Negative
scores are read as missing; valid categories begin at zero.

```{r fit}
d <- simulate_rasch(
  n_persons = 600,
  n_items = 12,
  model = "PCM",
  n_categories = 4,
  difficulty = c(-1.5, 1.5),
  disordered = "I04",
  dependence = list(pairs = list(c("I10", "I11")), strength = 1.3),
  dif = list(items = "I08", uniform = 0.8),
  n_groups = 3,
  seed = 17
)

fit <- rasch(d, model = "PCM", id = "id", factors = "group")
```

## Begin with the summary and reliability

The summary establishes whether estimation converged and gives the overall
item--trait interaction, fit-residual distributions, reliability, and the
package's qualitative assessment of the power of fit tests. It should be read
before individual item or person flags.

```{r overall-summary}
fit
fit_summary_table(fit)
```

The person separation index (PSI) compares the observed variance of person
locations with their mean error variance:

$$
\mathrm{PSI}=
\frac{\operatorname{Var}(\hat\theta)-
\operatorname{mean}\{\operatorname{SE}(\hat\theta)^2\}}
{\operatorname{Var}(\hat\theta)}.
$$

The implementation truncates negative values at zero. `fit$psi_noext` removes
persons with extreme scores and is useful when extremes inflate the observed
spread. The separation ratio and number of strata express the same information
on more interpretable scales.

```{r reliability}
fit$psi
fit$psi_noext
fit$power_of_fit
```

The power label is a screening judgement based on PSI, not a formal power
calculation. Low reliability weakens the ordering of persons and the formation
of distinct class intervals, so item--trait and DIF tests may fail to detect
real departures. Power also depends on sample size, test length, targeting,
category use, and trait spread. Conversely, a large sample can make a small
departure statistically significant. Fit residuals, effect sizes, plots, and
substantive importance remain necessary.

## Examine item estimates and fit

Item locations and their standard errors should be read alongside fit
residuals, mean-square statistics, and the Holm-adjusted item--trait
probabilities. A positive residual indicates more variation than expected; a
negative residual indicates responses that are more predictable than expected.
The conventional $\pm2.5$ residual band is a screening rule rather than a
separate hypothesis test.

```{r item-estimates}
fit$items[, c(
  "item", "location", "se", "fit_resid",
  "infit_ms", "outfit_ms", "p_adj"
)]
```

```{r item-fit-plot, fig.alt = "Item locations plotted against item fit residuals."}
plot_item_map(fit)
```

For a polytomous item, successive thresholds should normally increase on the
latent scale. Threshold disordering means that an intervening category is not
the most probable response over any part of the scale. It can reflect sparse
categories, indistinct category meanings, or a scoring order that is not
working as intended.

```{r thresholds}
fit$thresholds
ordered <- vapply(fit$thresholds_diag, function(x) x$ordered, logical(1))
disordered_items <- names(ordered)[!ordered]
disordered_items
fit$thresholds_diag[disordered_items]
```

```{r category-curves, fig.alt = "Category characteristic curves and observed category proportions for item I04."}
plot_ccc(fit, "I04", observed = TRUE)
```

Do not collapse categories from the ordering flag alone. Check category
frequencies, the category curves, the wording of the response options, and the
consequences for measurement. If categories are combined on substantive
grounds, refit the model and repeat the full diagnostic sequence.

## Examine person estimates and fit

Person fit addresses the consistency of each response pattern with the fitted
scale. The following table puts the largest absolute residuals first. Positive
residuals indicate unexpectedly erratic patterns; negative residuals indicate
patterns that are unusually predictable.

```{r person-estimates}
person_order <- order(abs(fit$person$fit_resid), decreasing = TRUE,
                      na.last = TRUE)
head(fit$person[person_order, c(
  "id", "group", "raw", "theta", "se", "extreme",
  "fit_resid", "infit_ms", "outfit_ms"
)], 10)
```

```{r person-fit-plot, fig.alt = "Person locations plotted against person fit residuals."}
plot_person_fit(fit)
```

An unexpected response pattern may reflect coding or data-entry errors,
careless responding, a secondary trait, or a genuine but unusual person. It is
not, by itself, a reason to remove the person. Fit residuals are unavailable
for extreme response patterns because those patterns do not provide an
interior location at which fit can be assessed.

## Examine targeting and information

Targeting concerns the match between the person distribution and the item
threshold distribution. The table reports their locations and spread, the
proportions of persons beyond the threshold range, and the principal
reliability indices.

```{r targeting-summary}
targeting_table(fit)
```

For a Rasch model, test information is the sum of the conditional response
variances:

$$
I(\theta)=\sum_i \operatorname{Var}(X_i\mid\theta),
\qquad
\operatorname{SE}(\hat\theta)\approx I(\theta)^{-1/2}.
$$

The person--item map below places person locations and item thresholds on the
same logit scale. The information curve shows where the test is most precise.

```{r targeting-map, fig.height = 5.2, fig.alt = "Person and item distributions with the test information curve on the common logit scale."}
plot_pimap(fit, information = TRUE)
```

```{r information-table}
head(test_information(fit))
```

The optional `WrightMap` package supplies a second form of Wright map. The
default retains one person panel and displays the estimated category
thresholds. Separate person or item panels should be requested only when they
answer a substantive question.

```{r wrightmap, fig.width = 8, fig.height = 5.2, fig.alt = "Wright map of person locations and item thresholds."}
if (requireNamespace("WrightMap", quietly = TRUE)) {
  wright_map(fit)
}
```

If `WrightMap` is not installed, install it with
`install.packages("WrightMap")` before running this chunk.

## Examine local and trait dependence

Local response dependence occurs when two responses remain associated after
conditioning on the latent trait. Yen's $Q3$ is the correlation between two
items' standardised residuals. Because raw $Q3$ values have a negative
baseline in a finite test, `q3_star` subtracts the average off-diagonal value.

```{r local-dependence}
q3 <- residual_correlations(fit)
q3$average
head(q3$pairs, 10)
```

```{r local-dependence-plot, fig.alt = "Heatmap of adjusted residual correlations between items."}
plot_resid_cor(fit)
```

There is no universal critical value for adjusted $Q3$ (Christensen,
Makransky and Horton 2017). Its size, the response process, and the content of
the item pair matter. When theory supports treating dependent items as one
superitem, `combine_items()` refits the complete calibration and
`spread_test()` compares its threshold spread with the binomial bound. The
bound does not apply to a superitem containing a polytomous component.

Trait dependence is examined through the residual components and by comparing
person estimates from opposed item subsets (Smith 2002). For subsets $A$ and
$B$, the person-level statistic is

$$
t_n=\frac{\hat\theta_{nA}-\hat\theta_{nB}}
{\sqrt{\operatorname{SE}(\hat\theta_{nA})^2+
       \operatorname{SE}(\hat\theta_{nB})^2}}.
$$

Under unidimensionality, about the nominated alpha level of these comparisons
should be significant. The exact binomial interval, the residual loadings,
and the score points in each subset are part of the result.

```{r trait-dependence}
dimensionality_test(fit)
```

```{r trait-dependence-plot, fig.alt = "Loadings of items on the first residual component."}
plot_pca(fit)
```

A quiet result from short subtests is inconclusive rather than evidence that a
secondary trait is absent. The returned caution records when either subset has
fewer score points than recommended for stable person estimates.

## Examine differential item functioning

Person factors must be nominated when the model is fitted. With a person
factor $G$ and trait class interval $C$, the residual model is

$$
z=\mu+G+C+G\mathbin{:}C+\varepsilon.
$$

The factor term tests uniform DIF; the interaction with class interval tests
non-uniform DIF. `dif_anova()` fits all nominated factors together, recognises
within-person factors in repeated designs, and applies Holm familywise
correction over the item-by-term tests.

```{r dif}
dif <- dif_anova(fit, sizes = TRUE)
subset(dif$summary, uniform_DIF | nonuniform_DIF)
```

```{r dif-plot, fig.alt = "Observed and expected item characteristic curves for item I08 by person group."}
plot_icc(fit, "I08", group = "group")
```

For a significant factor with more than two levels, the follow-up should
estimate the relevant differences in Rasch logits rather than rely on a
generic Tukey procedure. With `sizes = TRUE`, `dif_anova()` returns the
Holm-adjusted marginal pairwise comparisons for significant main effects and
difference-in-differences for significant interactions.

```{r dif-follow-up}
dif$posthoc
```

Statistical significance and practical magnitude answer different questions.
Any split must be supported by the response process and should be applied with
`resolve_dif()`, which refits the calibration and updates the item and person
estimates. The revised fit then goes through the same summary, fit, targeting,
dependence, and DIF sequence.

## Link calibrations through common items

Two separately fitted tests can be aligned through their common items. Drift
tests are meaningful only when the calibrations use independent sampling
units; state this rather than allowing the software to assume it.

```{r equating-example, eval = FALSE}
eq <- equate_tests(current_fit, reference_fit, independent = TRUE)
eq$table
```

A saved item bank needs more than marginal standard errors for drift
inference. Its joint item-location covariance carries the dependence created
by the fitted origin and is attached as `attr(bank, "cov_location")`, in bank
row order or named by item. Without that matrix, the package still estimates
the origin shift and reports the link descriptively. A bank intended to be
fixed may instead use zero standard errors. For polytomous items, include the
maximum score in a `max` column so that scoring-scale compatibility can be
checked.

## Report the final calibration

A report should state the model, scoring and missing-data rules, sample and
item exclusions, reliability, item and person fit, threshold functioning,
targeting, dependence, dimensionality, DIF, and the uncertainty of person
estimates. The score table supplies the complete-response conversion from raw
score to logits.

```{r reporting}
score_table(fit)
```

## References

Andrich, D. (1978). Relationships between the Thurstone and Rasch
approaches to item scaling. *Applied Psychological Measurement*, 2(3),
451--462.

Andrich, D., and Marais, I. (2019). *A Course in Rasch Measurement Theory:
Measuring in the Educational, Social and Health Sciences*. Springer.

Christensen, K. B., Makransky, G., and Horton, M. (2017). Critical values for
Yen's Q3: Identification of local dependence in the Rasch model using residual
correlations. *Applied Psychological Measurement*, 41(3), 178--194.

Holm, S. (1979). A simple sequentially rejective multiple test procedure.
*Scandinavian Journal of Statistics*, 6(2), 65--70.

Humphry, S. M. (2005). *Maintaining a Common Arbitrary Unit in Social
Measurement*. PhD thesis, Murdoch University.

Humphry, S. M., and Andrich, D. (2008). Understanding the unit in the
Rasch model. *Journal of Applied Measurement*, 9(3), 249--264.

Rasch, G. (1960). *Probabilistic Models for Some Intelligence and Attainment
Tests*. Copenhagen: Danish Institute for Educational Research. (Expanded
edition, 1980, Chicago: University of Chicago Press.)

Smith, E. V. Jr. (2002). Detecting and evaluating the impact of
multidimensionality using item fit statistics and principal component analysis
of residuals. *Journal of Applied Measurement*, 3(2), 205--231.

Tutz, G. (1986). Bradley-Terry-Luce models with an ordered response.
*Journal of Mathematical Psychology*, 30(3), 306--316.

Torres Irribarra, D., and Freund, R. (2025). *WrightMap: IRT item-person map
with ConQuest integration*. R package version 1.5.

Warm, T. A. (1989). Weighted likelihood estimation of ability in item
response theory. *Psychometrika*, 54(3), 427--450.

Yen, W. M. (1984). Effects of local item dependence on the fit and equating
performance of the three-parameter logistic model. *Applied Psychological
Measurement*, 8(2), 125--145.
